### 2019-07-22 manuscript.Rmd
Draft version 1

---
title: "Effects of PCSK9 Inhibition With Evolocumab on Lipoprotein Metabolism in subjects with elevated Lp(a)"
output: word_document
bibliography: /home/xiang/Documents/Bibliography/lpa.bib
csl: circulation.csl
---

Xiang Zhang, PhD$^1$,
Lotte C.A. Stiekema, MD$^2$,
Erik S.G. Stroes, PhD$^2$,
Albert K. Groen, PhD$^1$

1. Department of Experimental Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
2. Department of Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands

# Abstract

Background: 
Epidemiological studies show that subjects with elevated lipoprotein(a) [Lp(a)] have higher risk for developing cardiovascular diseases. 
Inhibition of proprotein convertase subtilisin/kexin type 9 (PCSK9) is one of the limited options to reduce Lp(a). 
In this study, we aimed to investigate the effects of PCSK9 inhibition with evolocumab on lipoprotein metabolism in subjects with elevated Lp(a).

Methods:
We performed metabolic profiling of plasma samples derived from 30 individuals (placebo N = 14; evolocumab N = 16). 
By applying Bayesian lognormal regression, we assessed the effects of evolocumab treatment on circulating Lp(a) as well as 14 lipoprotein subclasses. 
Our further regression analysis revealed the distribution of LDL particles with and without apolipoprotein(a) [apo(a)] before and after evolocumab treatment.

Results:
We identified that evolocumab treatment for 16 weeks resulted in 18.9% reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles. 
Moreover, we identified that evolocumab treatment not only reduced LDL particle concentrations, but also the distributions between LDL particles with and without apo(a).

Conclusions:
Based on our analysis, we suggest that the LDL particles with and without apo(a) tail may have different affinity for binding to the LDL receptor. 
Evolocumab-induced reduction of LDL particles may mainly correspond to the reduction of LDL particles without apo(a).

# Introduction

Lipoprotein(a) [Lp(a)] is an independent risk factor for cardiovascular diseases, and often considered as a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail produced by *LPA*. 
Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration. 
Proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody such as evolocumab has been shown to lower Lp(a) [@Raal2014; @Watts2017; @Shapiro2019].  
Evolocumab (also called AMG 145) is a fully human monoclonal antibody that targets PCSK9 and prevents it from binding to the LDL receptor. 
Although evolocumab-induced Lp(a) reduction was linked to enhanced LDL receptor mediated uptake and attenuated Lp(a) production, the mechanism of Lp(a) lowering by PCSK9 inhibition is unknown [@Romagnuolo2015; @Raal2016; @Watts2018]. 
In the current study, we aim to study the effects of PCSK9 inhibition with evolocumab on lipoprotein metabolism in subjects with elevated Lp(a). 
We decided to focus on lipoprotein metabolism, because genetic inhibition of PCSK9 and statin therapy were found to have similar metabolic effects, but differ in very-low-density lipoprotein (VLDL) metabolism [@Sliz2018]. 
So far, effects of PCSK9 inhibition with monoclonal antibody on lipoprotein metabolism were only assessed in healthy volunteers [@Reyes-Soffer2017]. 
The next step is to investigate the metabolic effects of PCSK9 inhibition on lipoprotein metabolism in subjects with high risk for cardiovascular diseases. 
To this end, we performed metabolic profiling of plasma samples derived from 30 individuals (placebo 14; evolocumab 16) with elevated Lp(a). 
We identified that despite evolocumab treatment for 16 weeks resulted in 18.9% reduction of Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles.   
Our further analysis revealed that evolocumab treatment not only reduced LDL particle concentrations, but also changed the distribution between LDL particles with and without apo(a).

# Methods

## Study design

This was a site-specific sub study of the ANITSCHKOW trial (NCT02729025) conducted at the Amsterdam UMC in The Netherlands between April 2016 and July 2017. 
The ANITSCHKOW trial was a phase 3b, multicenter, randomized, double-blind, placebo-controlled trial using subcutaneous injections of evolocumab 420mg Q4W for 16 weeks as investigational product [@Stiekema2018].
The study was conducted in accordance to the principles of the Declaration of Helsinki and the study protocol was approved by the ethic committee of the Amsterdam UMC.

Patients were found to be eligible for enrollment in the ANITSCHKOW trial when they were ≥ 50 years of age, had a fasting LDL-C of ≥ 2.6 mmol/L (100mg/dL), and an Lp(a) of ≥ 125nmol/L (50mg/dL). 
At time of randomization, study subjects who were on lipid lowering treatment had to be on a stable dosing regimen for at least 8 weeks. 
Exclusion criteria were among others, diabetes mellitus and a cardiovascular event within three months prior to randomization. 
The complete list of eligibility criteria was reported previously [@Stiekema2018].

## Biochemical measurements

Blood samples were obtained at time of randomization and sixteen weeks after treatment. 
Patients were fasting for ≥ 9 hours for both blood withdrawals. 
Lipoprotein(a) levels were measured using an isoform-independent immunoturbidometric assay (Polymedco, Cortlandt Manor, NY, USA) and reported in nmol/L.

## Metabolite quantification

Quantification of 14 lipoproteins was performed by using a high-throughput NMR metabolomics platform (Nightingale, Austria)
The following 14 lipoprotein subclasses were quantified: 
extremely large (average particle diameter >75 nm), 
very large (average particle diameter 64.0 nm), large (53.6 nm), medium (44.5 nm), small (36.8 nm) and very small VLDL (31.3 nm); 
intermediate density lipoprotein (IDL; 28.6 nm); 
three LDL subclasses, i.e. large (25.5 nm), medium (23.0 nm) and small LDL (18.7 nm); 
and four HDL subclasses, i.e. very large (14.3 nm), large (12.1 nm), medium (10.9 nm) and small HDL (8.7 nm).

## Statistical analysis

### Effects of evolocumab on Lp(a) and lipoprotein particles

The outcome variable ($y$) was concentration of Lp(a) or one of the 14 lipoprotein subclasses. 
Two predictor variables were used: 1) evolocumab treatment (evolocumab, $T_i$ = 1; placebo $T_i = 0$); 2) baseline Lp(a) or lipoprotein particle concentration ($x$).
The baseline concentrations were centered and scaled so that the mean is 0 and standard deviation is 1.

To assess metabolic effects of evolocumab treatment, we ran lognormal regression analysis. 
The lognormal distribution was chosen to model the outcome variable because its values were positive continuous. 
Due to missing observations in outcome variables, we applied the Bayesian approach to handle the missing data. 
There are two types of missing values, which are as follows: 
(1) when the concentration of a metabolite is below the limit of detection, 
or (2) when values were rejected by the automatic sample and measurement quality control procedure in the Nightingale pipeline. 
All the missing observations were assumed missing at random and treated as parameters.
Values were randomly drawn from a lognormal distribution. 
Regarding the missing values that were below the limit of detection, the imputed values were constrained between zero and the minimal observed value. 
We fitted the model by running Hamiltonian Markov Chain Monte Carlo in the program Stan (version 2.18.3).
We ran four Markov chains with 2000 iterations in each chain. 
Results were presented with the posterior mean of lognormal regression coefficient with 95% credible interval (CI). 
The lognormal regression coefficient can be interpreted as the expected proportional change in the metabolite concentration.

### Effects of evolocumab on composition of LDL particles

Because Lp(a) is often considered to comprise a LDL particle and covalently bound apo(a), we assumed that a pool of LDL particles is comprised by LDL particles with and without apo(a). 
In the Nightingale platform, three kinds of LDL particles (large, medium and small VLDL) were profiled. 
To identify the proportion of particles with apo(a) in each of these three pools, we ran Bayesian regression analysis. 
The outcome variable ($y$) was concentration of Lp(a). 
Three predictor variables were used: 1) Large LDL particle concentration $x_1$; 2) Medium LDL particle concentration ($x_2$); 3) Small LDL particle concentration ($x_3$). 
We excluded the intercept from our regression model and used the beta distribution $Beta(1,1)$ as the prior distribution for the regression coefficients. 
Because of the same motivation dealing with missing data, we applied the Bayesian approach. 
As mentioned above, we fitted the model by running four Markov chains with 2000 iterations in each chain. 
The regression coefficients are interpreted as proportion of LDL particles with apo(a) tail.

# Results

The current study included 30 patients (evolocumab, n = 16; placebo, n = 14). 
Baseline characteristics of the two groups were comparable and provided in Table 1.

## Effects on lipoprotein particles

Evolocumab treatment for 16 weeks resulted in mean 18.9% (95% credible interval: 7.51% to 29.2%) reduction in Lp(a) compared with placebo (Figure 1).

![Figure 1 Evolocumab treatment for 16 weeks resulted in mean (95% credible interval) reduction in Lp(a) compared to placebo group: 24.5% (14.0% to 35.5%)](./Treatment_effect_lpa.png)

To identify the corresponding effects of evolocumab on lipoprotein metabolism, we focused on the 14 lipoprotein subclasses. 
We observed that evolocumab treatment resulted in substantial reduction in particle concentration of extremely large (-2.09 [-4.25 -0.284]), very large (-2.87 [-5.53 -0.488]), large (-0.973 [-1.69, -0.350]), medium (-0.702 [-0.979, -0.413]), small (-0.492 [-0.610 -0.367]) and very small VLDL (-0.640 [-0.770 -0.515]) compared with placebo.
We also observed that compared with placebo evolocumab treatment resulted in particle concentration reduction in IDL (-0.765 [-0.930, -0.587]), large (-0.842 [-1.05 -0.657]), medium (-0.901 [-1.12, -0.698]) and small LDL (-0.810 [-1.00, -0.610]). 
In contrast to VLDL and LDL particles, we observed that evolocumab treatment resulted in increasing concentrations of medium HDL particle (0.119 [0.0366, 0.200]) (Figure 2).

![Figure 2 Mean difference between evolocumab and palacebo group, adjusting for pre-treatment lipoprotein particle concentrations. Circles represent the posterior means. Lines refer to the 95% credible intervals](./Treatment_effect_lipoproteins.png)

## Effects on LDL particle compositions

Lp(a) is thought to be comprised of a LDL particle and a covalently bound apo(a). 
In the Nightingale metabolomics platform, three LDL particles, including large, medium and small LDL, were profiled. 
At baseline, we identified that on average 43.9% (95% credible interval [0.0102%, 89.3%]) large LDL, 40.9% ([0.00807%, 88.6%]) medium LDL particles and 50.5% ([5.39% 97.3%]) small LDL particles were with apo(a) (Figure 3). 
The correlations of LDL-apo(a) proportion between these three LDL particles were -0.4 (large and medium LDL), -0.62 (large and small LDL) and -0.42 (medium and small LDL).

![Figure 3 Posterior mean of LDL-apo(a) proportion in large, medium and small LDL particles. Each dot (in total 4000 dots) is a mean proportion of LDL particles with apolipoprotein(a) sampled from the posterior distribtuions. The marginal distribution of the posterior mean was on the diagnal. The pearson correlation coefficients were shown](./Lpa_LDL_baseline.png)

After treatment of evolocumab, on average 76.5% (41.5%, 100%) large LDL particles were with apo(a). 
In the pool of medium LDL particles, on average 69.2% (25.5%, 100%) of them were with apo(a). 
Regarding small LDL particles, on average 75.8% (39.3%, 100%) of them were with apo(a).

![Figure 4 Posterior mean of LDL-apo(a) proportion in large, medium and small LDL particles. Each dot (in total 4000 dots) is a mean proportion of LDL particles with apolipoprotein(a) sampled from the posterior distribtuions. The marginal distribution of the posterior mean was on the diagnal. The pearson correlation coefficients were shown](./Lpa_LDL_visit2.png)

# Discussion

In the current study, evolocumab treatment for 16 weeks resulted in on average 18.9% reduction in Lp(a) compared with placebo.
Our detailed lipoprotein profiling revealed that treatment of evolocumab substantially reduced concentrations of VLDL, IDL and LDL particles. 
Our observation is consistent with the previous study, in which evolocumab treatment not only reduced concentrations of LDL particles, but also substantially reduced triglyceride-rich lipoprotein particles [@Toth2018].
The next question we came up with is "why we observed large effect of evolocumab on VLDL, IDL and LDL particle concentrations but relatively small effect on Lp(a)?" 
To address the question, we analyzed the composition of LDL particles. 
Because Lp(a) is often thought to be comprised of a LDL particle and a covalently bound apo(a), we assumed that a pool of LDL particles contained LDL particles with and without apo(a) tail. 
At baseline, we identified that on average 40%-50% of large, medium and small LDL particles had apo(a) tail. 
After evolocumab treatment, we identified that the proportion of LDL with apo(a) in all three pools of LDL particles increased to about 70%-75%. 
The composition shifting to relatively more LDL particles with apo(a) indicated that the reduction of LDL particles may mainly correspond to the reduction of LDL particles without apo(a). 
Indeed Lp(a) needs to compete with LDL particles for binding to the LDL receptor, and is preferred for uptake by the LDL receptor when the LDL particle concentration is low [@Raal2016].  
We hypothesized that the LDL particles with and without apo(a) tail may have different affinity for binding to the LDL receptor. 
Besides LDL receptor, we acknowledge that VLDL receptor can also mediate Lp(a) catabolism [@Argraves1997]. 
In our analysis, we observed substantial reduction in VLDL particle concentrations after treatment of evolocumab. 
Since PCSK9 can induce the degradation of LDL receptor as well as VLDL receptor, evolocumab treatment must enhance activity of both receptors [@Poirier2008]. 
Interestingly, we observed that most of patients receiving evolocumab ended in very low concentration (below the detection limit) of extremely large and very large VLDL particles, suggesting enhancement of VLDL receptor activity.

## Strengths and limitations

Strengths and limitations of our study warrant consideration. 
Our study provided first insight into relationship between Lp(a) concentrations and NMR LDL particle concentrations. 
One limitation of our study is that due to small sample size, our uncertainty of LDL particle composition estimation is large.

# Authors

# Acknowledgements

# Sources of Funding

# Disclosures

# Affiliations

# References

### 2019-07-22 manuscript.Rmd
Update manuscript.Rmd
Draft version 1 with modified introduction

@@
 Background: 
-Epidemiological studies show that subjects with elevated lipoprotein(a) [Lp(a)] have higher risk for developing cardiovascular diseases. 
+Epidemiological studies show that subjects with elevated lipoprotein(a) [Lp(a)] have a higher risk for developing cardiovascular diseases. 
 Inhibition of proprotein convertase subtilisin/kexin type 9 (PCSK9) is one of the limited options to reduce Lp(a). 
@@
 Proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody such as evolocumab has been shown to lower Lp(a) [@Raal2014; @Watts2017; @Shapiro2019].
+PCSK9 binds to hepatic low density lipoprotein (LDL) receptors and mediates their degradation in lysosome.
 Evolocumab (also called AMG 145) is a fully human monoclonal antibody that targets PCSK9 and prevents it from binding to the LDL receptor. 
 Although evolocumab-induced Lp(a) reduction was linked to enhanced LDL receptor mediated uptake and attenuated Lp(a) production, the mechanism of Lp(a) lowering by PCSK9 inhibition is unknown [@Romagnuolo2015; @Raal2016; @Watts2018].
-In the current study, we aim to study the effects of PCSK9 inhibition with evolocumab on lipoprotein metabolism in subjects with elevated Lp(a). 
-We decided to focus on lipoprotein metabolism, because genetic inhibition of PCSK9 and statin therapy were found to have similar metabolic effects, but differ in very-low-density lipoprotein (VLDL) metabolism [@Sliz2018].
-So far, effects of PCSK9 inhibition with monoclonal antibody on lipoprotein metabolism were only assessed in healthy volunteers [@Reyes-Soffer2017]. 
-The next step is to investigate the metabolic effects of PCSK9 inhibition on lipoprotein metabolism in subjects with high risk for cardiovascular diseases. 
-To this end, we performed metabolic profiling of plasma samples derived from 30 individuals (placebo 14; evolocumab 16) with elevated Lp(a). 
+As shown in a recent meta-analysis, useage of statins leads to increasing of Lp(a) concentration [@Tsimikas2019]. 
+More interestingly, decreasing Lp(a) concentration was seen in homozygous familial hypercholesterolaemia (FH) patients after treatment of evolocumab [@Raal2015]. 
+These observations suggested existence of non-LDL receptor mediated catabolism of Lp(a). 
+In the current study, we aim to explore the alternative pathways beyond LDL receptor-mediated catabolism of Lp(a).
+To this end, we studied the metabolic effects of PCSK9 inhibition with evolocumab in 30 subjects (placebo 14; evolocumab 16) with elevated Lp(a). 
 We identified that despite evolocumab treatment for 16 weeks resulted in 18.9% reduction of Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles.   
@@
 
-Patients were found to be eligible for enrollment in the ANITSCHKOW trial when they were ≥ 50 years of age, had a fasting LDL-C of ≥ 2.6 mmol/L (100mg/dL), and an Lp(a) of ≥ 125nmol/L (50mg/dL). 
+Patients were found to be eligible for enrollment in the ANITSCHKOW trial when they were $\geq$ 50 years of age, had a fasting LDL-C of $\geq$ 2.6 mmol/L (100mg/dL), and an Lp(a) of $\geq$ 125nmol/L (50mg/dL).
 At time of randomization, study subjects who were on lipid lowering treatment had to be on a stable dosing regimen for at least 8 weeks. 
@@
 Blood samples were obtained at time of randomization and sixteen weeks after treatment. 
-Patients were fasting for ≥ 9 hours for both blood withdrawals. 
+Patients were fasting for $\geq$ 9 hours for both blood withdrawals. 
 Lipoprotein(a) levels were measured using an isoform-independent immunoturbidometric assay (Polymedco, Cortlandt Manor, NY, USA) and reported in nmol/L.

### 2019-10-04 manuscript.Rmd
add new affiliation
new figure for LPa-apo(a)

@@
 
-Xiang Zhang, PhD$^1$,
+Xiang Zhang, PhD$^1,^2$,
 Lotte C.A. Stiekema, MD$^2$,
@@
 1. Department of Experimental Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
-2. Department of Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
+2. Human and Animal Physiology, Wageningen University, De Elst 1, 6708 WD Wageningen, The Netherlands
+3. Department of Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
 
@@
 
-![Figure 1 Evolocumab treatment for 16 weeks resulted in mean (95% credible interval) reduction in Lp(a) compared to placebo group: 24.5% (14.0% to 35.5%)](./Treatment_effect_lpa.png)
+![Figure 1 Evolocumab treatment for 16 weeks resulted in mean (95% credible interval) reduction in Lp(a) compared to placebo group: 18.9% (7.51% to 29.2%)](./Treatment_effect_lpa.png)
 
@@
 In the Nightingale metabolomics platform, three LDL particles, including large, medium and small LDL, were profiled. 
-At baseline, we identified that on average 43.9% (95% credible interval [0.0102%, 89.3%]) large LDL, 40.9% ([0.00807%, 88.6%]) medium LDL particles and 50.5% ([5.39% 97.3%]) small LDL particles were with apo(a) (Figure 3).
-The correlations of LDL-apo(a) proportion between these three LDL particles were -0.4 (large and medium LDL), -0.62 (large and small LDL) and -0.42 (medium and small LDL).
+At baseline, we identified that on average 56% (95% credible interval [9%, 100%]) large LDL, 50% ([3%, 97%]) medium LDL particles and 53% ([7% 99%]) small LDL particles were with apo(a) (Figure 3).
 
-![Figure 3 Posterior mean of LDL-apo(a) proportion in large, medium and small LDL particles.
-Each dot (in total 4000 dots) is a mean proportion of LDL particles with apolipoprotein(a) sampled from the posterior distribtuions.
-The marginal distribution of the posterior mean was on the diagnal.
-The pearson correlation coefficients were shown](./Lpa_LDL_baseline.png)
+After treatment of evolocumab, on average 70% (28%, 100%) large LDL particles were with apo(a). 
+In the pool of medium LDL particles, on average 61% (11%, 99%) of them were with apo(a). 
+Regarding small LDL particles, on average 68% (24%, 100%) of them were with apo(a).
 
-After treatment of evolocumab, on average 76.5% (41.5%, 100%) large LDL particles were with apo(a). 
-In the pool of medium LDL particles, on average 69.2% (25.5%, 100%) of them were with apo(a). 
-Regarding small LDL particles, on average 75.8% (39.3%, 100%) of them were with apo(a).
-
-![Figure 4 Posterior mean of LDL-apo(a) proportion in large, medium and small LDL particles.
-Each dot (in total 4000 dots) is a mean proportion of LDL particles with apolipoprotein(a) sampled from the posterior distribtuions.
-The marginal distribution of the posterior mean was on the diagnal.
-The pearson correlation coefficients were shown](./Lpa_LDL_visit2.png)
+![Figure 3 Proportions of particles with and without apo(a) in large, medium and small LDL pools.
+The sizes of pie charts were scaled to the average particle concentrations.](./Lpa_LDL_composition.png)
 

### 2019-10-04 manuscript.Rmd
add discussion on apo(a) composition

@@
 Although evolocumab-induced Lp(a) reduction was linked to enhanced LDL receptor mediated uptake and attenuated Lp(a) production, the mechanism of Lp(a) lowering by PCSK9 inhibition is unknown [@Romagnuolo2015; @Raal2016; @Watts2018].
-As shown in a recent meta-analysis, useage of statins leads to increasing of Lp(a) concentration [@Tsimikas2019]. 
+As shown in a recent meta-analysis, usage of statins leads to increasing of Lp(a) concentration [@Tsimikas2019]. 
 More interestingly, decreasing Lp(a) concentration was seen in homozygous familial hypercholesterolaemia (FH) patients after treatment of evolocumab [@Raal2015]. 
@@
 Because Lp(a) is often thought to be comprised of a LDL particle and a covalently bound apo(a), we assumed that a pool of LDL particles contained LDL particles with and without apo(a) tail. 
-At baseline, we identified that on average 40%-50% of large, medium and small LDL particles had apo(a) tail. 
-After evolocumab treatment, we identified that the proportion of LDL with apo(a) in all three pools of LDL particles increased to about 70%-75%. 
-The composition shifting to relatively more LDL particles with apo(a) indicated that the reduction of LDL particles may mainly correspond to the reduction of LDL particles without apo(a). 
+At baseline, we identified that on average about 50% of large, medium and small LDL particles had apo(a) tail. 
+Evolocumab treatment not only decreased concentrations of LDL particles, but also resulted in 10-20% increasing in LDL particles with apo(a). 
+The composition shifting to relatively more LDL particles with apo(a) indicated that LDL particles with apo(a) were less preferably cleared from the circulation compared to their counterparts without apo(a).
 Indeed Lp(a) needs to compete with LDL particles for binding to the LDL receptor, and is preferred for uptake by the LDL receptor when the LDL particle concentration is low [@Raal2016].  

### 2019-10-07 manuscript.Rmd
Modified abstract

@@
 Inhibition of proprotein convertase subtilisin/kexin type 9 (PCSK9) is one of the limited options to reduce Lp(a). 
-In this study, we aimed to investigate the effects of PCSK9 inhibition with evolocumab on lipoprotein metabolism in subjects with elevated Lp(a).
+In this study, we aimed to investigate metabolic effects of PCSK9 inhibition with evolocumab in subjects with elevated Lp(a).
 
@@
 We performed metabolic profiling of plasma samples derived from 30 individuals (placebo N = 14; evolocumab N = 16). 
-By applying Bayesian lognormal regression, we assessed the effects of evolocumab treatment on circulating Lp(a) as well as 14 lipoprotein subclasses. 
-Our further regression analysis revealed the distribution of LDL particles with and without apolipoprotein(a) [apo(a)] before and after evolocumab treatment.
+By applying Bayesian lognormal regression, we assessed the effects of evolocumab treatment on circulating metabolites such as lipoproteins and amino acids. 
+Further hierarchical modeling, we also studied effects of evolocumab treatment on proportions of LDL particles with and without apolipoprotein(a) [apo(a)].
 
@@
 We identified that evolocumab treatment for 16 weeks resulted in 18.9% reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles. 
-Moreover, we identified that evolocumab treatment not only reduced LDL particle concentrations, but also the distributions between LDL particles with and without apo(a).
+Moreover, we observed that evolocumab treatment tended to increase the proportion of LDL particles with apo(a).
 
 Conclusions:
-Based on our analysis, we suggest that the LDL particles with and without apo(a) tail may have different affinity for binding to the LDL receptor. 
-Evolocumab-induced reduction of LDL particles may mainly correspond to the reduction of LDL particles without apo(a).
+Based on our analysis, we showed that evolocumab treatment had large effects on lipoprotein metabolism.
 

### 2019-10-09 manuscript.Rmd
introduction for draft version 3

@@
 
-Lipoprotein(a) [Lp(a)] is an independent risk factor for cardiovascular diseases, and often considered as a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail produced by *LPA*.
-Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration. 
-Proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody such as evolocumab has been shown to lower Lp(a) [@Raal2014; @Watts2017; @Shapiro2019].
-PCSK9 binds to hepatic low density lipoprotein (LDL) receptors and mediates their degradation in lysosome.
-Evolocumab (also called AMG 145) is a fully human monoclonal antibody that targets PCSK9 and prevents it from binding to the LDL receptor. 
-Although evolocumab-induced Lp(a) reduction was linked to enhanced LDL receptor mediated uptake and attenuated Lp(a) production, the mechanism of Lp(a) lowering by PCSK9 inhibition is unknown [@Romagnuolo2015; @Raal2016; @Watts2018].
-As shown in a recent meta-analysis, usage of statins leads to increasing of Lp(a) concentration [@Tsimikas2019]. 
-More interestingly, decreasing Lp(a) concentration was seen in homozygous familial hypercholesterolaemia (FH) patients after treatment of evolocumab [@Raal2015]. 
-These observations suggested existence of non-LDL receptor mediated catabolism of Lp(a). 
-In the current study, we aim to explore the alternative pathways beyond LDL receptor-mediated catabolism of Lp(a).
-To this end, we studied the metabolic effects of PCSK9 inhibition with evolocumab in 30 subjects (placebo 14; evolocumab 16) with elevated Lp(a). 
-We identified that despite evolocumab treatment for 16 weeks resulted in 18.9% reduction of Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles.   
-Our further analysis revealed that evolocumab treatment not only reduced LDL particle concentrations, but also changed the distribution between LDL particles with and without apo(a).
+lipoprotein(a) [Lp(a)], synthesized in liver, is a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail. 
+Similar to other lipoproteins, Lp(a) is cleared from the circulation primarily by the liver [@Cain2005]. 
+Epidemiological studies showed that subjects with elevated Lp(a) had increasing risk of developing cardiovascular disease [@ODonoghue2019]. 
+Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration [@VanCapelleveen2016]. 
+One of the agents that are capable to lower circulating Lp(a) levels is proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody, which prevents extracellular PCSK9 from binding to the LDL receptor, and results in supraphysiological abundance of LDL receptors [@Raal2014; @Raal2016].
+The excess number of LDL receptors on the surface of hepatocytes were hypothesized to be responsible for the enhanced Lp(a) catabolism [@Romagnuolo2015; @Raal2016]. 
+However, the above hypothesis does not concur with the effect of statins therapy, which significantly increases hepatic LDL receptor abundances but also increases plasma Lp(a) levels [@Tsimikas2019]. 
+The same hypothesis was also challenged by the observation that LDL receptor knock out mice had similar Lp(a) catabolism as the wild type mice [@Cain2005]. 
+Therefore, what roles LDL-receptor plays in Lp(a) catabolism has not been elucidated and remains a matter of debate [@Reyes-Soffer2017; @Watts2018; @Stoekenbroek2018]. 
+In addition to the LDL receptor, very low density lipoprotein (VLDL) receptor and scavenger receptor class B type I (SR-BI) were also found to act as potential receptor for Lp(a), indicating more than one pathway is likely to be involved in Lp(a) reduction by PCKS9 inhibition [@Argraves1997; @Yang2013].
+As a cost-effective approach, nuclear magnetic resonance (NMR) metabolomics simultaneously quantifies hundreds of circulating biomarkers and provides insights on multiple pathways [@Soininen2009].
+In fact, NMR metabolomics has been used to study metabolic effects that were associated with  *LPA* loci, as well as genetic inhibition of PCSK9 [@Kettunen2016; @Sliz2018]. 
+However, genetic inhibition of PCSK9 is different from inhibition with monoclonal antibody, and there is lack of studies on metabolic effects of PCSK9 inhibition with monoclonal antibody. 
+In the current study, we aim to study metabolic changes corresponding to Lp(a) reduction in subjects with elevated Lp(a) after treatment of PCSK9 monoclonal antibody (evolocumab).
+To this end, we performed NMR metabolomics of plasma samples derived from 30 healthy participants with elevated Lp(a) levels (placebo 14; evolocumab 16). 
+Since Lp(a) and LDL significantly overlap in density, protein and lipid composition, LDL measurements made by current technologies contain the content of Lp(a) [@Yeang2015]. 
+To address this issue, in this study we developed a statistical model to distinguish these two lipoproteins.
 

### 2019-10-10 manuscript.Rmd
Method section for draft version 3

@@
 
-## Biochemical measurements
+## Metabolite quantification
 
@@
 Patients were fasting for $\geq$ 9 hours for both blood withdrawals.
+
 Lipoprotein(a) levels were measured using an isoform-independent immunoturbidometric assay (Polymedco, Cortlandt Manor, NY, USA) and reported in nmol/L.
 
-## Metabolite quantification
-
-Quantification of 14 lipoproteins was performed by using a high-throughput NMR metabolomics platform (Nightingale, Austria)
-The following 14 lipoprotein subclasses were quantified: 
+Quantification of 225 metabolic measures was performed by using a high-throughput NMR metabolomics platform (Nightingale health, Finland) [@Soininen2009]. 
+The 225 metabolic measures contain around 150 primary concentrations as well as ratios that were derived from the primary concentrations. 
+In this study, we focused on primary concentrations of circulating metabolites that cover multiple metabolic pathways including lipoproteins, fatty acids as well as amino acids and glycolysis intermediates.
+The following 14 lipoprotein subclasses and their lipid compositions were quantified: 
 extremely large (average particle diameter >75 nm), 
@@
 and four HDL subclasses, i.e. very large (14.3 nm), large (12.1 nm), medium (10.9 nm) and small HDL (8.7 nm).
+The complete list of the 225 metabolic measures can be found at  https://nightingalehealth.com/biomarkers
 
@@
 
-### Effects of evolocumab on Lp(a) and lipoprotein particles
+### Metabolic effects of PCSK9 inhibition with evolocumab
 
-The outcome variable ($y$) was concentration of Lp(a) or one of the 14 lipoprotein subclasses. 
-Two predictor variables were used: 1) evolocumab treatment (evolocumab, $T_i$ = 1; placebo $T_i = 0$); 2) baseline Lp(a) or lipoprotein particle concentration ($x$).
+To assess the effect of evolocumab on levels of a circulating metabolite, we ran lognormal regression analysis. 
+The outcome variable ($y$) was concentration of a metabolite. 
+Two predictor variables were in the regression model: 1) evolocumab treatment (evolocumab, $T_i$ = 1; placebo $T_i = 0$); 
+2) metabolite concentration at time of randomization ($x$).
 The baseline concentrations were centered and scaled so that the mean is 0 and standard deviation is 1.
-
-To assess metabolic effects of evolocumab treatment, we ran lognormal regression analysis. 
 The lognormal distribution was chosen to model the outcome variable because its values were positive continuous. 
@@
 
-### Effects of evolocumab on composition of LDL particles
+### Separating Lp(a) content from LDL measures
 
-Because Lp(a) is often considered to comprise a LDL particle and covalently bound apo(a), we assumed that a pool of LDL particles is comprised by LDL particles with and without apo(a). 
-In the Nightingale platform, three kinds of LDL particles (large, medium and small VLDL) were profiled. 
-To identify the proportion of particles with apo(a) in each of these three pools, we ran Bayesian regression analysis. 
-The outcome variable ($y$) was concentration of Lp(a). 
-Three predictor variables were used: 1) Large LDL particle concentration $x_1$; 2) Medium LDL particle concentration ($x_2$); 3) Small LDL particle concentration ($x_3$). 
-We excluded the intercept from our regression model and used the beta distribution $Beta(1,1)$ as the prior distribution for the regression coefficients. 
-Because of the same motivation dealing with missing data, we applied the Bayesian approach. 
-As mentioned above, we fitted the model by running four Markov chains with 2000 iterations in each chain. 
-The regression coefficients are interpreted as proportion of LDL particles with apo(a) tail.
+Since Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition, LDL measures made by current technologies contain the content of Lp(a). 
+In this study, we developed a statistical model to address this issue. 
+Our statistical model was inspired by the biological relationship between Lp(a) and LDL. 
+Lp(a) is often considered to comprise a LDL particle and covalently bound apo(a). 
+Based on the above knowledge, we assumed that a pool of LDL particles is comprised by LDL particles with and without apo(a). 
+The outcome variable ($y$) was the particle concentration of Lp(a), which was measured by using the isoform-independent immunoturbidometric assay. 
+Again we used the lognormal distribution as likelihood function for the Lp(a) concentration. 
+Four predictor variables were used: 
+1) Large LDL particle concentration $\mathsf{LLDL}$; 
+2) Medium LDL particle concentration ($\mathsf{MLDL}$);
+3) Small LDL particle concentration ($\mathsf{SLDL}$);
+4) Whether the measurement is made at time of randomization or sixteen weeks after treatment (16 weeks after, $V = 1$).  
+These particle concentrations were measured by using the NMR metabolomics. 
+Because lipid-lowering medication such as statins, ezetimibe as well as evolocumab has causal effects on these LDL particle concentrations, we included three covariates (statin usage: $S = 1$; Ezetimibe usage $E = 1$; Evolocumab usage $P = 1$).
+The mathematical details of the model was given below:
+$$\mathsf{Lp_a}[i] \sim Lognormal(\mu[i],\sigma)$$
+$$exp(\mu[i]) = \theta_L \times exp(\mu_{LLDL}[i]) + \theta_M \times exp(\mu_{MLDL}[i]) + \theta_S \times exp(\mu_{SLDL}[i])$$
+$$LLDL[i] \sim Lognormal(\mu_{LLDL[i]}, \sigma_{L})$$
+$$MLDL[i] \sim Lognormal(\mu_{MLDL[i]}, \sigma_{M})$$
+$$SLDL[i] \sim Lognormal(\mu_{SLDL[i]}, \sigma_{S})$$
+$$exp(\mu_{LLDL}[i]) = exp(\alpha_L) + \beta_{S,L} \times S[i] + \beta_{E,L} \times E[i] + (\beta_L + \beta_{P,L} \times P[i]) \times V[i]$$
+$$exp(\mu_{MLDL}[i]) = exp(\alpha_M) + \beta_{S,M} \times S[i] + \beta_{E,M} \times E[i] + (\beta_M + \beta_{P,M} \times P[i]) \times V[i]$$
+$$exp(\mu_{SLDL}[i]) = exp(\alpha_S) + \beta_{S,S} \times S[i] + \beta_{E,S} \times E[i] + (\beta_S + \beta_{P,S} \times P[i]) \times V[i]$$
+
+Regarding $\theta_L$, $\theta_M$ and $\theta_S$, we set the beta distribution $Beta(1,1)$ as their prior distributions, since these regression coefficients are interpreted as proportion of LDL particles with apo(a) tail.
 

### 2019-10-10 manuscript.Rmd
result section for draft version 3

@@
 
-## Effects on lipoprotein particles
+## Metabolic effects of PCSK9 inhibition with evolocumab
 
-Evolocumab treatment for 16 weeks resulted in mean 18.9% (95% credible interval: 7.51% to 29.2%) reduction in Lp(a) compared with placebo (Figure 1).
+Evolocumab treatment for 16 weeks resulted in mean 18.9% (95% credible interval: 7.51% to 29.2%) reduction in Lp(a) compared with placebo.
 
-![Figure 1 Evolocumab treatment for 16 weeks resulted in mean (95% credible interval) reduction in Lp(a) compared to placebo group: 18.9% (7.51% to 29.2%)](./Treatment_effect_lpa.png)
+To identify the metabolic effects corresponding to the Lp(a) reduction by PCSK9 inhibition, we performed NMR metabolomics covering metabolic pathways such as lipoproteins, fatty acids, amino acids and glycolysis.
+We observed that evolocumab treatment resulted in substantial reduction in particle concentration of extremely large (81% [51% 100%]), very large (89% [69% 100%]), large (61% [36%, 86%]), medium (50% [36%, 63%]), small (39% [31% 46%]) and very small VLDL (47% [41% 54%]) compared with the placebo group.
+We also observed that evolocumab treatment resulted in particle concentration reduction in IDL (53% [45%, 61%]), large (57% [49% 66%]), medium (59% [50%, 6%]) and small LDL (55% [47%, 64%]) compared with the placebo group.
+In addition, we observed that evolocumab treatment resulted in decreased concentrations of very large HDL particles (25% [4% 46%]), and increased concentrations of medium HDL particles (13% [4% 23%]) (Figure 1).
 
-To identify the corresponding effects of evolocumab on lipoprotein metabolism, we focused on the 14 lipoprotein subclasses. 
-We observed that evolocumab treatment resulted in substantial reduction in particle concentration of extremely large (-2.09 [-4.25 -0.284]), very large (-2.87 [-5.53 -0.488]), large (-0.973 [-1.69, -0.350]), medium (-0.702 [-0.979, -0.413]), small (-0.492 [-0.610 -0.367]) and very small VLDL (-0.640 [-0.770 -0.515]) compared with placebo.
-We also observed that compared with placebo evolocumab treatment resulted in particle concentration reduction in IDL (-0.765 [-0.930, -0.587]), large (-0.842 [-1.05 -0.657]), medium (-0.901 [-1.12, -0.698]) and small LDL (-0.810 [-1.00, -0.610]).
-In contrast to VLDL and LDL particles, we observed that evolocumab treatment resulted in increasing concentrations of medium HDL particle (0.119 [0.0366, 0.200]) (Figure 2).
-
-![Figure 2 Mean difference between evolocumab and palacebo group, adjusting for pre-treatment lipoprotein particle concentrations.
+![Figure 1 Mean difference between evolocumab and palacebo group, adjusting for pre-treatment lipoprotein particle concentrations.
 Circles represent the posterior means.
@@
 
-## Effects on LDL particle compositions
+Similar to the lipoprotein particle concentration profiles, we observed that evolocumab treatment resulted in substantial reduction of esterified cholesterol (CE), triglyceride (TG), free cholesterol (FC) and phospholipid (PL) in VLDL, IDL and LDL as well as very large HDL, compared with the placebo group (Figure 2).
+
+![Figure 2 Mean difference in lipoprotein lipid compositions between evolocumab and placebo group, adjusting for pre-treatment lipoprotein lipid concentrations.
+Circles represent the posterior means.
+Lines refer to the 95% credible intervals.
+CE: esterified cholesterol; FC: free cholesterol; PL: phospholipids; TG: triglyceride.](./Treatment_effect_lipoprotein_lipid.png)
+
+The NMR metabolomics also quantified the fatty acid content in the lipoproteins. 
+We observed that treatment of evolocumab resulted in 30% [24% 36%] reduction in total fatty acids, with the largest effect on docosahexaenoic acid (DHA 40% [24% 55%]) (Figure 3).
+
+![Figure 2 Mean difference in fatty acid concentrations between evolocumab and placebo group, adjusting for pre-treatment fatty acid concentrations.
+Circles represent the posterior means.
+Lines refer to the 95% credible intervals.](./Treatment_effect_FFAs.png)
+
+We observed no difference in concentrations of other metabolites covering amino acids, fluid balance, glycolysis and ketone bodies.
+
+## Separating Lp(a) content from the LDL measures
 
 Lp(a) is thought to be comprised of a LDL particle and a covalently bound apo(a). 
-In the Nightingale metabolomics platform, three LDL particles, including large, medium and small LDL, were profiled. 
-At baseline, we identified that on average 56% (95% credible interval [9%, 100%]) large LDL, 50% ([3%, 97%]) medium LDL particles and 53% ([7% 99%]) small LDL particles were with apo(a) (Figure 3).
+Because Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition, LDL measures contain the content of Lp(a). 
+At baseline, we identified that on average 58% ([8%, 100%]) large LDL, 49% ([0.2%, 95%]) medium LDL particles and 55% ([7% 100%]) small LDL particles were with apo(a) (Figure 5).
 
-After treatment of evolocumab, on average 70% (28%, 100%) large LDL particles were with apo(a). 
-In the pool of medium LDL particles, on average 61% (11%, 99%) of them were with apo(a). 
-Regarding small LDL particles, on average 68% (24%, 100%) of them were with apo(a).
+Sixteen weeks after treatment of evolocumab, on average 74% (28%, 100%) large LDL particles were with apo(a).  
+The average proportions of particles with apo(a) in the pool of medium and small LDL were 65% (12%, 100%) and 72% (24%, 100%), respectively (Figure 5).
 
-![Figure 3 Proportions of particles with and without apo(a) in large, medium and small LDL pools.
+![Figure 5 Proportions of particles with and without apo(a) in large, medium and small LDL pools.
 The sizes of pie charts were scaled to the average particle concentrations.](./Lpa_LDL_composition.png)

### 2019-10-10 manuscript.Rmd
discussion for draft version 3

@@
 
-Evolocumab treatment for 16 weeks resulted in mean 18.9% (95% credible interval: 7.51% to 29.2%) reduction in Lp(a) compared with placebo.
+Evolocumab treatment for 16 weeks resulted in mean 17% (95% credible interval: 8% to 26%) reduction in Lp(a) compared with placebo.
 
@@
 
-In the current study, evolocumab treatment for 16 weeks resulted in on average 18.9% reduction in Lp(a) compared with placebo.
-Our detailed lipoprotein profiling revealed that treatment of evolocumab substantially reduced concentrations of VLDL, IDL and LDL particles. 
+The major finding of this work is that evolocumab treatment in patients with high Lp(a) massively decreases large VLDL in addition to the well known effect on LDL. 
+In the current work, evolocumab treatment resulted in on average 17% reduction in plasma Lp(a) concentration compared with placebo. 
+Our detailed metabolomic profiling revealed that Lp(a) reduction by PCSK9 inhibition corresponded to substantial reduction in concentrations of VLDL, IDL and LDL particles as well as their lipid and fatty acid contents.
 Our observation is consistent with the previous study, in which evolocumab treatment not only reduced concentrations of LDL particles, but also substantially reduced triglyceride-rich lipoprotein particles [@Toth2018].
-The next question we came up with is "why we observed large effect of evolocumab on VLDL, IDL and LDL particle concentrations but relatively small effect on Lp(a)?" 
-To address the question, we analyzed the composition of LDL particles. 
-Because Lp(a) is often thought to be comprised of a LDL particle and a covalently bound apo(a), we assumed that a pool of LDL particles contained LDL particles with and without apo(a) tail. 
+Since a PCSK9 monoclonal antibody such as evolocumab reduces the degradation of the LDL receptor as well as the LDL receptor, the very low concentration of VLDL particles after the treatment may be ascribed to the increased umber of VLDL receptors [@Poirier2008].
+The upregulation of VLDL receptors can enhance uptake of VLDLs but also Lp(a) from the circulation [@Argraves1997]. 
+In addition, our analysis also showed that treatment of evolocumab lowered concentrations of very large HDL particles. 
+Very large HDL is cholesterol enriched and mainly captured by the SR-B1 present in the liver [@Acton1996]. 
+Similar to the VLDL receptor, SR-B1 has also been shown to bind Lp(a), and its upregulation will probably enhance the Lp(a) catabolism [@Yang2013].
+
+In our analysis, we observed a large discordance in Lp(a) and LDL reduction in response to evolocumab. 
+This discordance was suggested by a recent study to be related to alternative pathways of Lp(a) reduction beyond LDL receptor mediated clearance [@Shapiro2019]. 
+Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition. 
+As a result, current technologies including the NMR metabolomics used in this study cannot distinguish these two lipoproteins, and include the content of Lp(a) in the LDL measurements. 
+We think that it is important to separate the Lp(a) content from the LDL measurements in order to better understand the discordance between these two lipoproteins in response to evolocumab. 
+To this end, we developed a statistical model based on the assumption that Lp(a) was comprised of a LDL particle and a covalently bound apo(a). 
 At baseline, we identified that on average about 50% of large, medium and small LDL particles had apo(a) tail. 
-Evolocumab treatment not only decreased concentrations of LDL particles, but also resulted in 10-20% increasing in LDL particles with apo(a). 
-The composition shifting to relatively more LDL particles with apo(a) indicated that LDL particles with apo(a) were less preferably cleared from the circulation compared to their counterparts without apo(a).
-Indeed Lp(a) needs to compete with LDL particles for binding to the LDL receptor, and is preferred for uptake by the LDL receptor when the LDL particle concentration is low [@Raal2016].  
-We hypothesized that the LDL particles with and without apo(a) tail may have different affinity for binding to the LDL receptor. 
-Besides LDL receptor, we acknowledge that VLDL receptor can also mediate Lp(a) catabolism [@Argraves1997]. 
-In our analysis, we observed substantial reduction in VLDL particle concentrations after treatment of evolocumab. 
-Since PCSK9 can induce the degradation of LDL receptor as well as VLDL receptor, evolocumab treatment must enhance activity of both receptors [@Poirier2008]. 
-Interestingly, we observed that most of patients receiving evolocumab ended in very low concentration (below the detection limit) of extremely large and very large VLDL particles, suggesting enhancement of VLDL receptor activity.
+After evolocumab treatment, on average about 60%-70% of large, medium and small LDL particles having apo(a) tail. 
+The relatively more LDL particles with apo(a) indicated that LDL particles with apo(a) were less preferably cleared from the circulation compared to their counterparts without apo(a). 
+Indeed a recent study showed that LDL without apo(a) was cleared faster than Lp(a), which was found to compete with LDL particles for binding to the LDL receptor [@Viney2018; @Raal2016].
 
@@
 Strengths and limitations of our study warrant consideration. 
-Our study provided first insight into relationship between Lp(a) concentrations and NMR LDL particle concentrations. 
-One limitation of our study is that due to small sample size, our uncertainty of LDL particle composition estimation is large.
+Our study provided the first NMR metabolomics data for a PCSK9 inhibition trial. 
+The detailed metabolomic profiling revealed metabolic effects of the PCSK9 inhibitor, and improved our understanding of this lipid-lowering therapy. 
+Moreover, in this study we developed a statistical method to separate Lp(a) content from the LDL measurements. 
+However, due to the heterogeneity of Lp(a) and limited number of participants, our estimation of the proportions of LDL with apo(a) had very large uncertainty. 
+Applying our statistical model with a larger number of participants as well as validation in the laboratory are warranted in the future studies.
+
+In conclusion, our NMR metabolomic profiling revealed that Lp(a) reduction by PCSK9 inhibition corresponded to substantial reduction in VLDL, IDL and LDL particles as well as their lipid contents. 
+Our analysis also suggested that LDL with and without apo(a) had different preference to be cleared from the circulation.
 

### 2019-10-10 manuscript.Rmd
draft version 3

@@
 ---
-title: "Effects of PCSK9 Inhibition With Evolocumab on Lipoprotein Metabolism in subjects with elevated Lp(a)"
+title: "Metabolic Effects of PCSK9 Inhibition With Evolocumab in subjects with elevated Lp(a)"
 output: word_document
@@
 By applying Bayesian lognormal regression, we assessed the effects of evolocumab treatment on circulating metabolites such as lipoproteins and amino acids. 
-Further hierarchical modeling, we also studied effects of evolocumab treatment on proportions of LDL particles with and without apolipoprotein(a) [apo(a)].
+In order to separate Lp(a) content from the LDL measurements, we developed a statistical model which can be used to estimate the proportions of LDL particles with and without apolipoprotein(a) [apo(a)].
 
 Results:
-We identified that evolocumab treatment for 16 weeks resulted in 18.9% reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles. 
+We identified that evolocumab treatment for 16 weeks resulted in 18.9% reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
 Moreover, we observed that evolocumab treatment tended to increase the proportion of LDL particles with apo(a).
@@
 Conclusions:
-Based on our analysis, we showed that evolocumab treatment had large effects on lipoprotein metabolism.
+Our analysis suggested that alternative pathways beyond LDL receptor mediated clearance may be involved in Lp(a) reduction by PCSK9 inhibition.
+Moreover, LDL with and without apo(a) tended to have different preference to be cleared from the circulation.
 

### 2019-10-10 manuscript.Rmd
fix the figure number for draft version 3

@@
 
-![Figure 2 Mean difference in fatty acid concentrations between evolocumab and placebo group, adjusting for pre-treatment fatty acid concentrations.
+![Figure 3 Mean difference in fatty acid concentrations between evolocumab and placebo group, adjusting for pre-treatment fatty acid concentrations.
 Circles represent the posterior means.
@@
 Because Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition, LDL measures contain the content of Lp(a). 
-At baseline, we identified that on average 58% ([8%, 100%]) large LDL, 49% ([0.2%, 95%]) medium LDL particles and 55% ([7% 100%]) small LDL particles were with apo(a) (Figure 5).
+At baseline, we identified that on average 58% ([8%, 100%]) large LDL, 49% ([0.2%, 95%]) medium LDL particles and 55% ([7% 100%]) small LDL particles were with apo(a) (Figure 4).
 
 Sixteen weeks after treatment of evolocumab, on average 74% (28%, 100%) large LDL particles were with apo(a).  
-The average proportions of particles with apo(a) in the pool of medium and small LDL were 65% (12%, 100%) and 72% (24%, 100%), respectively (Figure 5).
+The average proportions of particles with apo(a) in the pool of medium and small LDL were 65% (12%, 100%) and 72% (24%, 100%), respectively (Figure 4).
 
-![Figure 5 Proportions of particles with and without apo(a) in large, medium and small LDL pools.
+![Figure 4 Proportions of particles with and without apo(a) in large, medium and small LDL pools.
 The sizes of pie charts were scaled to the average particle concentrations.](./Lpa_LDL_composition.png)

### 2019-10-10 manuscript.Rmd
fix abstract for the draft version 3

@@
 Results:
-We identified that evolocumab treatment for 16 weeks resulted in 18.9% reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
+We identified that evolocumab treatment for 16 weeks resulted in 17% (95% credible interval: 8% to 26%) reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
 Moreover, we observed that evolocumab treatment tended to increase the proportion of LDL particles with apo(a).
@@
 Our observation is consistent with the previous study, in which evolocumab treatment not only reduced concentrations of LDL particles, but also substantially reduced triglyceride-rich lipoprotein particles [@Toth2018].
-Since a PCSK9 monoclonal antibody such as evolocumab reduces the degradation of the LDL receptor as well as the LDL receptor, the very low concentration of VLDL particles after the treatment may be ascribed to the increased umber of VLDL receptors [@Poirier2008].
+Since a PCSK9 monoclonal antibody such as evolocumab reduces the degradation of the LDL receptor as well as the VLDL receptor, the very low concentration of VLDL particles after the treatment may be ascribed to the increased number of VLDL receptors [@Poirier2008].
 The upregulation of VLDL receptors can enhance uptake of VLDLs but also Lp(a) from the circulation [@Argraves1997]. 

### 2019-10-30 manuscript.Rmd
Replace the text on regression model for mixture model

@@
 By applying Bayesian lognormal regression, we assessed the effects of evolocumab treatment on circulating metabolites such as lipoproteins and amino acids. 
-In order to separate Lp(a) content from the LDL measurements, we developed a statistical model which can be used to estimate the proportions of LDL particles with and without apolipoprotein(a) [apo(a)].
+In order to separate Lp(a) content from the LDL measurements, we applied a Gaussian mixture model to estimate the proportions of LDL particles with and without apolipoprotein(a) [apo(a)].
 
@@
 We identified that evolocumab treatment for 16 weeks resulted in 17% (95% credible interval: 8% to 26%) reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
-Moreover, we observed that evolocumab treatment tended to increase the proportion of LDL particles with apo(a).
+We also identified that about 24% [0.2% 47%] of large LDL particles were with apo(a). 
+We did not observe changes in the proportion of LDL particles with apo(a) after treatment of evolocumab.
 
@@
 Our analysis suggested that alternative pathways beyond LDL receptor mediated clearance may be involved in Lp(a) reduction by PCSK9 inhibition.
-Moreover, LDL with and without apo(a) tended to have different preference to be cleared from the circulation.
 
@@
 Since Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition, LDL measures made by current technologies contain the content of Lp(a). 
-In this study, we developed a statistical model to address this issue. 
+In this study, we applied the Gaussian mixture model to address this issue. 
 Our statistical model was inspired by the biological relationship between Lp(a) and LDL. 
 Lp(a) is often considered to comprise a LDL particle and covalently bound apo(a). 
-Based on the above knowledge, we assumed that a pool of LDL particles is comprised by LDL particles with and without apo(a). 
-The outcome variable ($y$) was the particle concentration of Lp(a), which was measured by using the isoform-independent immunoturbidometric assay. 
-Again we used the lognormal distribution as likelihood function for the Lp(a) concentration. 
-Four predictor variables were used: 
-1) Large LDL particle concentration $\mathsf{LLDL}$; 
-2) Medium LDL particle concentration ($\mathsf{MLDL}$);
-3) Small LDL particle concentration ($\mathsf{SLDL}$);
-4) Whether the measurement is made at time of randomization or sixteen weeks after treatment (16 weeks after, $V = 1$).  
-These particle concentrations were measured by using the NMR metabolomics. 
+The diameter of Lp(a) is about 25 nm [@BERG1963]. 
+Based on the above knowledge, we assumed that the pool of large LDL particle (average diameter 25.5 nm) is a mixture of LDL particles with and without apo(a). 
+The outcome variable ($y$) was the particle concentration of large LDL, and described by a mixture of two normal distributions with a proportion parameter $\lambda$.
+
+$$P(y|\lambda, \mu, \sigma) = \prod_{i = 1}^{N}(\lambda \times N(\mu_1,\sigma_1) + (1 - \lambda) \times N(\mu_2, \sigma_2))$$
+
+The parameter $\mu_1$ and $\sigma_1$ were estimated based on the particle concentration of Lp(a), which was measured by using the isoform-independent immunoturbidometric assay. 
+The parameter $\mu_2$ was modeled as $\mu_2[i] = \alpha_L + \beta_{S,L} \times S[i] + \beta_{E,L} \times E[i] + (\beta_L + \beta_{P,L} \times P[i]) \times V[i]$.
+
+There were four covariates $V$, $S$, $E$ and $P$. 
+$V = 0$ referred the measurement was made at time of randomization, whereas $V = 1$ refereed to sixteen weeks after treatment. 
 Because lipid-lowering medication such as statins, ezetimibe as well as evolocumab has causal effects on these LDL particle concentrations, we included three covariates (statin usage: $S = 1$; Ezetimibe usage $E = 1$; Evolocumab usage $P = 1$).
-The mathematical details of the model was given below:
-$$\mathsf{Lp_a}[i] \sim Lognormal(\mu[i],\sigma)$$
-$$exp(\mu[i]) = \theta_L \times exp(\mu_{LLDL}[i]) + \theta_M \times exp(\mu_{MLDL}[i]) + \theta_S \times exp(\mu_{SLDL}[i])$$
-$$LLDL[i] \sim Lognormal(\mu_{LLDL[i]}, \sigma_{L})$$
-$$MLDL[i] \sim Lognormal(\mu_{MLDL[i]}, \sigma_{M})$$
-$$SLDL[i] \sim Lognormal(\mu_{SLDL[i]}, \sigma_{S})$$
-$$exp(\mu_{LLDL}[i]) = exp(\alpha_L) + \beta_{S,L} \times S[i] + \beta_{E,L} \times E[i] + (\beta_L + \beta_{P,L} \times P[i]) \times V[i]$$
-$$exp(\mu_{MLDL}[i]) = exp(\alpha_M) + \beta_{S,M} \times S[i] + \beta_{E,M} \times E[i] + (\beta_M + \beta_{P,M} \times P[i]) \times V[i]$$
-$$exp(\mu_{SLDL}[i]) = exp(\alpha_S) + \beta_{S,S} \times S[i] + \beta_{E,S} \times E[i] + (\beta_S + \beta_{P,S} \times P[i]) \times V[i]$$
-
-Regarding $\theta_L$, $\theta_M$ and $\theta_S$, we set the beta distribution $Beta(1,1)$ as their prior distributions, since these regression coefficients are interpreted as proportion of LDL particles with apo(a) tail.
+Regarding parameter $\lambda$,  proportion of LDL particles with apo(a) tail, we set the beta distribution $Beta(1,1)$ as its prior distributions.
 
@@
 Because Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition, LDL measures contain the content of Lp(a). 
-At baseline, we identified that on average 58% ([8%, 100%]) large LDL, 49% ([0.2%, 95%]) medium LDL particles and 55% ([7% 100%]) small LDL particles were with apo(a) (Figure 4).
+In this study, we focused on the large LDL particle, because its diameter (25.5 nm) is close to the diameter of Lp(a). 
+At baseline, we identified that on average 24% ([0.2%, 47%]) large LDL particles were with apo(a) (Figure 4).
 
-Sixteen weeks after treatment of evolocumab, on average 74% (28%, 100%) large LDL particles were with apo(a).  
-The average proportions of particles with apo(a) in the pool of medium and small LDL were 65% (12%, 100%) and 72% (24%, 100%), respectively (Figure 4).
-
-![Figure 4 Proportions of particles with and without apo(a) in large, medium and small LDL pools.
-The sizes of pie charts were scaled to the average particle concentrations.](./Lpa_LDL_composition.png)
+Sixteen weeks after treatment of evolocumab, on average 12% (0.02%, 36%) large LDL particles were with apo(a) in the placebo group. 
+In the treatment group, the average proportion of large LDL particles with apo(a) was 9% (0.0008%, 31%).
+We did not observe that the proportions of large LDL particles with apo(a) were different between the treatment and placebo group (-3% [-36% 23%]).
 
@@
 We think that it is important to separate the Lp(a) content from the LDL measurements in order to better understand the discordance between these two lipoproteins in response to evolocumab. 
-To this end, we developed a statistical model based on the assumption that Lp(a) was comprised of a LDL particle and a covalently bound apo(a). 
-At baseline, we identified that on average about 50% of large, medium and small LDL particles had apo(a) tail. 
-After evolocumab treatment, on average about 60%-70% of large, medium and small LDL particles having apo(a) tail. 
-The relatively more LDL particles with apo(a) indicated that LDL particles with apo(a) were less preferably cleared from the circulation compared to their counterparts without apo(a). 
-Indeed a recent study showed that LDL without apo(a) was cleared faster than Lp(a), which was found to compete with LDL particles for binding to the LDL receptor [@Viney2018; @Raal2016].
+To this end, we applied a Gaussian mixture model based on the assumption that Lp(a) was comprised of a LDL particle and a covalently bound apo(a). 
+We identified that on average about 24% of large LDL particle had apo(a) tail. 
+We did not observe changes in proportions of large LDL particle with apo(a) tail due to evolocumab treatment.
 
@@
 In conclusion, our NMR metabolomic profiling revealed that Lp(a) reduction by PCSK9 inhibition corresponded to substantial reduction in VLDL, IDL and LDL particles as well as their lipid contents. 
-Our analysis also suggested that LDL with and without apo(a) had different preference to be cleared from the circulation.
+Our analysis suggested that alternative pathways beyond LDL receptor mediated clearance may be involved in Lp(a) reduction by PCSK9 inhibition.
 

### 2019-12-13 manuscript.Rmd
story Lp(a) and TG metabolism

@@
 ---
-title: "Metabolic Effects of PCSK9 Inhibition With Evolocumab in subjects with elevated Lp(a)"
+title: "Metabolic Effects of PCSK9 Inhibition With Evolocumab in Subjects with Elevated Lp(a)"
 output: word_document
 bibliography: /home/xiang/Documents/Bibliography/lpa.bib
-csl: circulation.csl
+csl: journal-of-lipid-research.csl
 ---
@@
 Epidemiological studies show that subjects with elevated lipoprotein(a) [Lp(a)] have a higher risk for developing cardiovascular diseases. 
-Inhibition of proprotein convertase subtilisin/kexin type 9 (PCSK9) is one of the limited options to reduce Lp(a). 
-In this study, we aimed to investigate metabolic effects of PCSK9 inhibition with evolocumab in subjects with elevated Lp(a).
+Inhibition of proprotein convertase subtilisin/kexin type 9 (PCSK9) is one of the limited options to lower Lp(a). 
+However, effects of PCSK9 inhibition with evolocumab on circulating metabolites such as lipoprotein subclasses, amino acids and fatty acids remain to be characterized.
 
 Methods:
-We performed metabolic profiling of plasma samples derived from 30 individuals (placebo N = 14; evolocumab N = 16). 
-By applying Bayesian lognormal regression, we assessed the effects of evolocumab treatment on circulating metabolites such as lipoproteins and amino acids. 
-In order to separate Lp(a) content from the LDL measurements, we applied a Gaussian mixture model to estimate the proportions of LDL particles with and without apolipoprotein(a) [apo(a)].
+In this study, we performed nuclear magnetic resonance (NMR) metabolomics on plasma samples derived from 30 individuals (placebo N = 14; evolocumab N = 16). 
+We assessed the effects of evolocumab treatment on circulating metabolites by running lognormal regression analyses. 
+Subsequently, we assessed the interrelationship between Lp(a) and 14 lipoprotein subclasses in response to treatment of evolocumab, by running multilevel multivariate regression analyses.
 
 Results:
-We identified that evolocumab treatment for 16 weeks resulted in 17% (95% credible interval: 8% to 26%) reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
-We also identified that about 24% [0.2% 47%] of large LDL particles were with apo(a). 
-We did not observe changes in the proportion of LDL particles with apo(a) after treatment of evolocumab.
+On average, evolocumab treatment for 16 weeks resulted in a 17% (95% credible interval: 8% to 26%) reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
+Although, we did not observe correlations between Lp(a) lowering and reduction in other 14 lipoprotein subclasses, we identified that increasing concentrations of baseline Lp(a) were associated with larger reduction in triglyceride-rich VLDL particles after evolocumab treatment.
 
 Conclusions:
-Our analysis suggested that alternative pathways beyond LDL receptor mediated clearance may be involved in Lp(a) reduction by PCSK9 inhibition.
+Inhibition of PCSK9 with evolocumab massively decreased VLDL in addition to the well known effect on LDL. 
+The substantial reduction in VLDLs after evolocumab treatment indicates enhanced clearance of VLDL particles in patients with high Lp(a).
 
@@
 
-lipoprotein(a) [Lp(a)], synthesized in liver, is a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail. 
+lipoprotein(a) [Lp(a)], synthesized in the liver, is a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail. 
 Similar to other lipoproteins, Lp(a) is cleared from the circulation primarily by the liver [@Cain2005]. 
 Epidemiological studies showed that subjects with elevated Lp(a) had increasing risk of developing cardiovascular disease [@ODonoghue2019]. 
-Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration [@VanCapelleveen2016]. 
+The biological mechanism underlying the cardiovascular risk of Lp(a) remains to be elucidated. 
+One mechanism is that Lp(a) may induce macrophage apoptosis and plaque necrosis through activation of a pathway involving CD36 (Cluster of differentiation 36) and TLR2 (Toll-like receptor 2) [@Seimon2010].
+Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce Lp(a) in plasma [@VanCapelleveen2016]. 
 One of the agents that are capable to lower circulating Lp(a) levels is proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody, which prevents extracellular PCSK9 from binding to the LDL receptor, and results in supraphysiological abundance of LDL receptors [@Raal2014; @Raal2016].
 The excess number of LDL receptors on the surface of hepatocytes were hypothesized to be responsible for the enhanced Lp(a) catabolism [@Romagnuolo2015; @Raal2016]. 
-However, the above hypothesis does not concur with the effect of statins therapy, which significantly increases hepatic LDL receptor abundances but also increases plasma Lp(a) levels [@Tsimikas2019]. 
-The same hypothesis was also challenged by the observation that LDL receptor knock out mice had similar Lp(a) catabolism as the wild type mice [@Cain2005]. 
-Therefore, what roles LDL-receptor plays in Lp(a) catabolism has not been elucidated and remains a matter of debate [@Reyes-Soffer2017; @Watts2018; @Stoekenbroek2018]. 
-In addition to the LDL receptor, very low density lipoprotein (VLDL) receptor and scavenger receptor class B type I (SR-BI) were also found to act as potential receptor for Lp(a), indicating more than one pathway is likely to be involved in Lp(a) reduction by PCKS9 inhibition [@Argraves1997; @Yang2013].
-As a cost-effective approach, nuclear magnetic resonance (NMR) metabolomics simultaneously quantifies hundreds of circulating biomarkers and provides insights on multiple pathways [@Soininen2009].
-In fact, NMR metabolomics has been used to study metabolic effects that were associated with  *LPA* loci, as well as genetic inhibition of PCSK9 [@Kettunen2016; @Sliz2018]. 
-However, genetic inhibition of PCSK9 is different from inhibition with monoclonal antibody, and there is lack of studies on metabolic effects of PCSK9 inhibition with monoclonal antibody. 
-In the current study, we aim to study metabolic changes corresponding to Lp(a) reduction in subjects with elevated Lp(a) after treatment of PCSK9 monoclonal antibody (evolocumab).
-To this end, we performed NMR metabolomics of plasma samples derived from 30 healthy participants with elevated Lp(a) levels (placebo 14; evolocumab 16). 
-Since Lp(a) and LDL significantly overlap in density, protein and lipid composition, LDL measurements made by current technologies contain the content of Lp(a) [@Yeang2015]. 
-To address this issue, in this study we developed a statistical model to distinguish these two lipoproteins.
+However, the above hypothesis does not concur with the effect of statin therapy, which significantly increases hepatic LDL receptor abundances but also increases plasma Lp(a) levels [@Tsimikas2019]. 
+The same hypothesis was also challenged by the observation that LDL receptor knock out mice had similar Lp(a) catabolism compared to wild type mice [@Cain2005]. 
+Hence, the exact role of the LDL-receptor in Lp(a) catabolism has not been elucidated and remains a matter of debate [@Reyes-Soffer2017; @Watts2018; @Stoekenbroek2018]. 
+The LDL receptor is not the only target of PCSK9. 
+PCSK9 also induces the degradation of the VLDL (very low density lipoprotein) receptor, ApoER2 (apolipoprotein E receptor 2), and CD36 [@Poirier2008; @Demers2015]. 
+Interestingly, the VLDL receptor and CD36 are receptors for triglyceride-rich lipoproteins (TRLs) as well as Lp(a), indicating a possible interrelationship between Lp(a) and TRL metabolism [@Argraves1997; @Seimon2010; @Dijk2018].
+The metabolic link between Lp(a) and TRL metabolism was confirmed by a recent NMR (nuclear magnetic resonance) metabolomics study, in which a Lp(a)-raising genotype was identified to associate with decreasing concentrations of triglyceride-rich VLDL particles [@Kettunen2016].
+However, the mechanism underlying this inverse relationship is unknown. 
+As a cost-effective approach, NMR metabolomics simultaneously quantifies hundreds of circulating biomarkers and provides insights on multiple metabolic pathways [@Soininen2009].
+A NMR metabolomics study showed that genetic inhibition of PCSK9 and statin therapy had similar effects on circulating metabolites [@Sliz2018]. 
+However, genetic inhibition of PCSK9 is different from inhibition with monoclonal antibody since genetic inhibition includes intracellular functions of PCSK9. 
+Furthermore, although PCSK9 monoclonal antibody is known for lowering LDL cholesterol and Lp(a), the relationship between Lp(a) lowering and metabolism of other lipoproteins has not been studied.  
+In the current study, we analyzed NMR metabolomics of plasma derived from 30 subjects with elevated Lp(a) (placebo 14; evolocumab 16). 
+We first ran lognormal regression analyses to identify metabolic changes corresponding to evolocumab treatment.  
+Next, we developed a multilevel multivariate model to assess the relationship between Lp(a) lowering and metabolism of 14 lipoprotein subclasses.
 
@@
 Patients were found to be eligible for enrollment in the ANITSCHKOW trial when they were $\geq$ 50 years of age, had a fasting LDL-C of $\geq$ 2.6 mmol/L (100mg/dL), and an Lp(a) of $\geq$ 125nmol/L (50mg/dL).
-At time of randomization, study subjects who were on lipid lowering treatment had to be on a stable dosing regimen for at least 8 weeks. 
+At the time of randomization, study subjects who were on lipid lowering treatment had to be on a stable dosing regimen for at least 8 weeks. 
 Exclusion criteria were among others, diabetes mellitus and a cardiovascular event within three months prior to randomization. 
@@
 
+## Biochemical measurements
+
+Blood samples were obtained at time of randomization and 16 weeks after treatment. Patients were fasting for $\geq$ 9 hours for both blood withdrawals. 
+Total cholesterol, high-density lipoprotein (HDL) cholesterol, triglycerides, and apolipoproteinB-100 (ApoB-100) were measured by commercially available kits at the Medpace core lab (Medpace Reference Laboratories, Leuven, Belgium).
+Low-density lipoprotein (LDL) cholesterol was calculated using the Friedewald formula. 
+Ultracentrifugation-determined LDL-C was measured and reported if calculated LDL cholesterol was below 40mg/dL, or triglycerides were above 400mg/dL. 
+Lipoprotein(a) levels were measured using an isoform-independent immunoturbidometric assay (Polymedco, Cortlandt Manor, NY, USA) and reported in nmol/L.
+
 ## Metabolite quantification
-
-Blood samples were obtained at time of randomization and sixteen weeks after treatment. 
-Patients were fasting for $\geq$ 9 hours for both blood withdrawals.
-
-Lipoprotein(a) levels were measured using an isoform-independent immunoturbidometric assay (Polymedco, Cortlandt Manor, NY, USA) and reported in nmol/L.
 
@@
 To assess the effect of evolocumab on levels of a circulating metabolite, we ran lognormal regression analysis. 
-The outcome variable ($y$) was concentration of a metabolite. 
+The outcome variable ($y$) was the concentration of a metabolite. 
 Two predictor variables were in the regression model: 1) evolocumab treatment (evolocumab, $T_i$ = 1; placebo $T_i = 0$); 
@@
 The lognormal distribution was chosen to model the outcome variable because its values were positive continuous. 
-Due to missing observations in outcome variables, we applied the Bayesian approach to handle the missing data. 
-There are two types of missing values, which are as follows: 
+Due to missing observations in outcome variables, we applied the Bayesian approach to handle missing data. 
+There were two types of missing values: 
 (1) when the concentration of a metabolite is below the limit of detection, 
@@
 We ran four Markov chains with 2000 iterations in each chain. 
-Results were presented with the posterior mean of lognormal regression coefficient with 95% credible interval (CI). 
-The lognormal regression coefficient can be interpreted as the expected proportional change in the metabolite concentration.
+Results were presented with the posterior mean with 95% credible interval (CI).
 
-### Separating Lp(a) content from LDL measures
+### Multilevel multivariate model
 
-Since Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition, LDL measures made by current technologies contain the content of Lp(a). 
-In this study, we applied the Gaussian mixture model to address this issue. 
-Our statistical model was inspired by the biological relationship between Lp(a) and LDL. 
-Lp(a) is often considered to comprise a LDL particle and covalently bound apo(a). 
-The diameter of Lp(a) is about 25 nm [@BERG1963]. 
-Based on the above knowledge, we assumed that the pool of large LDL particle (average diameter 25.5 nm) is a mixture of LDL particles with and without apo(a). 
-The outcome variable ($y$) was the particle concentration of large LDL, and described by a mixture of two normal distributions with a proportion parameter $\lambda$.
+To assess the relationship between Lp(a) lowering and metabolism of 14 lipoprotein subclasses, we developed a multilevel multivariate model. 
+The detailed model is available at https://github.com/XiangZhangSC/Anitschkow.
+In short, 16 subjects from the treatment group were used for this analysis, and each subject had two plasma samples at time of randomization and after treatment.   
+The outcome variable is a vector of 2, $[L_i, y_i]$, in which $L_i$ represents the Lp(a) concentration in sample $i$, and $y_i$ represents the particle concentration of a lipoprotein subclass in sample $i$.
+The logarithm of $[L_i, y_i]$ was modeled by a multivariate normal distribution, with parameters $\mu_L$ (a vector of 32), $\mu_y$ (a vector of 32) and $\Sigma$ (a 2-by-2 covariance matrix). 
+$\mu_L[i]$ represents the mean concentration of Lp(a) in sample $i$, and was modeled as 
+$$\mu_L[i] = \alpha_{L,subject[i]} + \beta_{L,subject[i]} \times V_2[i]$$ 
+$\alpha_L$ is a vector of 16, and represents the Lp(a) abundance at baseline. 
+$\beta_L$ is a vector of 16, and represents the effect of evolocumab treatment on Lp(a). 
+$V_2$ is a vector of 32 with 0s and 1s. 
+When $V_2[i] = 1$, it means that measurement was derived from the second visit (after treatment). 
+Otherwise, $V_2[i] = 0$. 
+Similarly, $\mu_y[i]$ represents the mean concentration of a lipoprotein subclass in sample $i$, and was modeled as 
+$$\mu_y[i] = \alpha_{y,subject[i]} + \beta_{y,subject[i]} \times V_2[i]$$ 
+$\alpha_y$ is a vector of 16, and represents the lipoprotein subclass abundance at baseline. 
+$\beta_y$ is a vector of 16, and represents the effect of evolocumab treatment on the lipoprotein subclass. 
+To assess the relationship between the Lp(a) lowering and the metabolism of a lipoprotein subclass, we used another multivariate normal distribution to model the subject-specific parameters. 
+$$[\alpha_L, \beta_L, \alpha_y, \beta_y][j] \sim MultivariateNormal([\theta_L, \gamma_L, \theta_y, \gamma_y], \Sigma_{subj})$$
 
-$$P(y|\lambda, \mu, \sigma) = \prod_{i = 1}^{N}(\lambda \times N(\mu_1,\sigma_1) + (1 - \lambda) \times N(\mu_2, \sigma_2))$$
-
-The parameter $\mu_1$ and $\sigma_1$ were estimated based on the particle concentration of Lp(a), which was measured by using the isoform-independent immunoturbidometric assay. 
-The parameter $\mu_2$ was modeled as $\mu_2[i] = \alpha_L + \beta_{S,L} \times S[i] + \beta_{E,L} \times E[i] + (\beta_L + \beta_{P,L} \times P[i]) \times V[i]$.
-
-There were four covariates $V$, $S$, $E$ and $P$. 
-$V = 0$ referred the measurement was made at time of randomization, whereas $V = 1$ refereed to sixteen weeks after treatment. 
-Because lipid-lowering medication such as statins, ezetimibe as well as evolocumab has causal effects on these LDL particle concentrations, we included three covariates (statin usage: $S = 1$; Ezetimibe usage $E = 1$; Evolocumab usage $P = 1$).
-Regarding parameter $\lambda$,  proportion of LDL particles with apo(a) tail, we set the beta distribution $Beta(1,1)$ as its prior distributions.
+$\theta_L$, $\gamma_L$, $\theta_y$, $\gamma_y$ and $\Sigma_{subj}$ (a 4-by-4 covariance matrix) are the hyper-parameters. 
+We fitted the model by running Hamiltonian Markov Chain Monte Carlo in the program Stan (version 2.18.3).
+We ran four Markov chains with 2000 iterations in each chain. 
+Results were presented with the posterior mean with 95% credible interval (CI).
 
@@
 The current study included 30 patients (evolocumab, n = 16; placebo, n = 14). 
-Baseline characteristics of the two groups were comparable and provided in Table 1.
+Baseline characteristics of the two groups were comparable and are provided in Table 1.
 
@@
 
-Evolocumab treatment for 16 weeks resulted in mean 17% (95% credible interval: 8% to 26%) reduction in Lp(a) compared with placebo.
+On average, evolocumab treatment for 16 weeks resulted in a 17% (95% credible interval: 8% to 26%) reduction in Lp(a), together with a concomitant 67% (57%, 76%) and 21% (6% 35%) reduction in LDL cholesterol and triglyceride, respectively.
 
-To identify the metabolic effects corresponding to the Lp(a) reduction by PCSK9 inhibition, we performed NMR metabolomics covering metabolic pathways such as lipoproteins, fatty acids, amino acids and glycolysis.
-We observed that evolocumab treatment resulted in substantial reduction in particle concentration of extremely large (81% [51% 100%]), very large (89% [69% 100%]), large (61% [36%, 86%]), medium (50% [36%, 63%]), small (39% [31% 46%]) and very small VLDL (47% [41% 54%]) compared with the placebo group.
-We also observed that evolocumab treatment resulted in particle concentration reduction in IDL (53% [45%, 61%]), large (57% [49% 66%]), medium (59% [50%, 6%]) and small LDL (55% [47%, 64%]) compared with the placebo group.
-In addition, we observed that evolocumab treatment resulted in decreased concentrations of very large HDL particles (25% [4% 46%]), and increased concentrations of medium HDL particles (13% [4% 23%]) (Figure 1).
+To identify the metabolic effects corresponding to PCSK9 inhibition with evolocumab, we performed NMR metabolomics covering metabolic pathways such as lipoprotein subclasses, fatty acids, amino acids and glycolysis.
+We observed that evolocumab treatment resulted in substantial reduction in particle concentration of extremely large (80% [48% 100%]), very large (90% [70% 100%]), large (60% [34%, 83%]), medium (50% [36%, 63%]), small (39% [32% 46%]) and very small VLDL (47% [40% 53%]).
+We also observed that evolocumab treatment resulted in particle concentration reduction in IDL (53% [45%, 60%]), large (56% [48% 65%]), medium (59% [50%, 67%]) and small LDL (55% [47%, 64%]). 
+In addition, we observed that evolocumab treatment resulted in decreased concentrations of very large HDL particles (24% [3% 45%]), and increased concentrations of medium HDL particles (13% [4% 23%]) (Figure 1).
 
@@
 
-Similar to the lipoprotein particle concentration profiles, we observed that evolocumab treatment resulted in substantial reduction of esterified cholesterol (CE), triglyceride (TG), free cholesterol (FC) and phospholipid (PL) in VLDL, IDL and LDL as well as very large HDL, compared with the placebo group (Figure 2).
+Similar to the lipoprotein particle concentration profiles, we observed that evolocumab treatment resulted in substantial reduction of esterified cholesterol (CE), triglyceride (TG), free cholesterol (FC) and phospholipid (PL) in VLDL, IDL and LDL as well as very large HDL (Figure 2).
 
@@
 
-We observed no difference in concentrations of other metabolites covering amino acids, fluid balance, glycolysis and ketone bodies.
+We observed no difference in concentrations of other metabolites including amino acids, fluid balance, glycolysis and ketone bodies.
 
-## Separating Lp(a) content from the LDL measures
+## Relationship between Lp(a) lowering and reduction in lipoprotein subclasses
 
-Lp(a) is thought to be comprised of a LDL particle and a covalently bound apo(a). 
-Because Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition, LDL measures contain the content of Lp(a). 
-In this study, we focused on the large LDL particle, because its diameter (25.5 nm) is close to the diameter of Lp(a). 
-At baseline, we identified that on average 24% ([0.2%, 47%]) large LDL particles were with apo(a) (Figure 4).
+To identify the relationship between Lp(a) lowering and reduction in lipoprotein subclasses, we developed a multilevel multivariate model. 
+We observed that Lp(a) lowering was not associated with reduction of the 14 lipoprotein subclasses (Figure 4).
 
-Sixteen weeks after treatment of evolocumab, on average 12% (0.02%, 36%) large LDL particles were with apo(a) in the placebo group. 
-In the treatment group, the average proportion of large LDL particles with apo(a) was 9% (0.0008%, 31%).
-We did not observe that the proportions of large LDL particles with apo(a) were different between the treatment and placebo group (-3% [-36% 23%]).
+![Figure 4 Relationship between Lp(a) lowering and reduction in 14 lipoprotein subclasses.
+Circles represent the posterior mean Pearson correlation coefficients.
+Lines refer to the 95% credible intervals.](./corr_lipoproteins_lpa.png)
+
+However, we identified that the reduction in medium-sized VLDL particles was associated with increasing baseline Lp(a) concentrations (Pearson correlation coefficient -0.5 [-0.8, -0.06]). 
+The correlations between baseline Lp(a) concentrations and reduction in other VLDL particles, including extremely large, very large, large, small and very small VLDLs, were -0.1 [-0.6, 0.3], -0.03 [-0.4, 0.4], -0.3 [-0.7, 0.2], -0.1 [-0.7, 0.6], and 0.2 [-0.3, 0.7], respectively.
+
+![Figure 5 Reduction of medium VLDL particles correlated with baseline lipoprotein(a) concentrations.
+Every circle represents the posterior mean reduction of medium VLDL particle concentration and the posterior mean of baseline lipoprotein(a) in a patient treated with evolocumab.
+The verticle and horizontal bar represents the 95% credible interval.
+The blue dashed line represented the average percentage (50%) change in medium VLDL particle after evolocumab treatment.](./Baseline_lpa_mVLDL.png)
 
@@
 
-The major finding of this work is that evolocumab treatment in patients with high Lp(a) massively decreases large VLDL in addition to the well known effect on LDL. 
-In the current work, evolocumab treatment resulted in on average 17% reduction in plasma Lp(a) concentration compared with placebo. 
-Our detailed metabolomic profiling revealed that Lp(a) reduction by PCSK9 inhibition corresponded to substantial reduction in concentrations of VLDL, IDL and LDL particles as well as their lipid and fatty acid contents.
-Our observation is consistent with the previous study, in which evolocumab treatment not only reduced concentrations of LDL particles, but also substantially reduced triglyceride-rich lipoprotein particles [@Toth2018].
-Since a PCSK9 monoclonal antibody such as evolocumab reduces the degradation of the LDL receptor as well as the VLDL receptor, the very low concentration of VLDL particles after the treatment may be ascribed to the increased number of VLDL receptors [@Poirier2008].
-The upregulation of VLDL receptors can enhance uptake of VLDLs but also Lp(a) from the circulation [@Argraves1997]. 
-In addition, our analysis also showed that treatment of evolocumab lowered concentrations of very large HDL particles. 
-Very large HDL is cholesterol enriched and mainly captured by the SR-B1 present in the liver [@Acton1996]. 
-Similar to the VLDL receptor, SR-B1 has also been shown to bind Lp(a), and its upregulation will probably enhance the Lp(a) catabolism [@Yang2013].
+The major finding of this work is that evolocumab treatment in patients with high Lp(a) massively decreases VLDLs in addition to the well known effect on LDLs. 
+In the current work, evolocumab treatment resulted in on average 17% reduction in plasma Lp(a) concentration, together with 67% and 21% reduction in LDL cholesterol and triglyceride. 
+These treatment effects were similar to the outcomes reported in the past, indicating that the subjects involved in the current study were representative for the entire ANITSCHKOW trail [@Stiekema2018].
 
-In our analysis, we observed a large discordance in Lp(a) and LDL reduction in response to evolocumab. 
-This discordance was suggested by a recent study to be related to alternative pathways of Lp(a) reduction beyond LDL receptor mediated clearance [@Shapiro2019]. 
-Lp(a) and LDL significantly overlap in density, as well as protein and lipid composition. 
-As a result, current technologies including the NMR metabolomics used in this study cannot distinguish these two lipoproteins, and include the content of Lp(a) in the LDL measurements. 
-We think that it is important to separate the Lp(a) content from the LDL measurements in order to better understand the discordance between these two lipoproteins in response to evolocumab. 
-To this end, we applied a Gaussian mixture model based on the assumption that Lp(a) was comprised of a LDL particle and a covalently bound apo(a). 
-We identified that on average about 24% of large LDL particle had apo(a) tail. 
-We did not observe changes in proportions of large LDL particle with apo(a) tail due to evolocumab treatment.
+Our detailed metabolomic profiling revealed that PCSK9 inhibition with evolocumab resulted in a substantial reduction in concentrations of VLDL, IDL and LDL particles as well as their lipid and fatty acid contents.
+We did not observe changes in other circulating metabolites such as amino acids and ketone bodies. 
+In general, PCSK9 inhibition with evolocumab resulted in similar metabolic changes as genetic inhibition of *PCSK9* and statin therapy [@Wurtz2016; @Sliz2018]. 
+However, in contrast to genetic inhibition of *PCSK9* and statin therapy that had much larger effects on LDL particles than VLDL particles, we observed that evolocumab treatment induced a more substantial reduction in VLDL particles compared to LDL particles.
+Our multivariate analysis showed that Lp(a) lowering did not correlate with reduction in any of the 14 lipoprotein subclasses, suggesting alternative pathways in Lp(a) catabolism, such as scavenger receptor class B type I (SR-BI) [@Yang2013] and plasminogen receptor [@Sharma2017].
+
+Interestingly, we identified that subjects with higher baseline Lp(a) showed a larger reduction in medium VLDL particle concentrations after treatment with evolocumab. 
+We observed that subjects with higher baseline Lp(a) also showed tendency to have a larger reduction in extremely large and large VLDL particle concentrations after treatment with evolocumab. 
+However, those tendencies had large uncertainty due to considerable number of missing observations. 
+The percentage of missing observations in extremely large and large were 69% and 47%, respectively. 
+Similarly, since the very large VLDL particle had the largest percentage of missing data (72%) among all the 14 lipoprotein subclasses, we simply had not enough data to detect the relationship between its reduction and baseline Lp(a).
+In contrast, the percentage of missing observations in medium VLDL was 9%, allowing us to assess the relationship between its reduction and baseline Lp(a) levels. 
+On the other hand, we did not observe similar relationship between baseline Lp(a) and reduction in small and very small VLDL particles, although the percentages of missing observations were small for these two lipoprotein subclasses (6% and 3%), suggesting that Lp(a) may not influence catabolism of these two triglyceride-poor VLDL particles.
+Our observation is in line with the association between the Lp(a)-raising genotype and decreasing concentrations of extremely large, very large, large and medium VLDL particles, but not small and very small VLDL particles [@Kettunen2016].
+Lp(a) particles can bind noncovalently to triglyceride-rich lipoproteins (TRL), forming a Lp(a)-TRL complex which was suggested to facilitate receptor-mediated uptake [@McConathy1994; @Gaubatz2001]. 
+This process may be particularly important in our subjects with high Lp(a), since they may have a large amount of the Lp(a)-TRL complex. 
+Together with the supraphysiological abundance of lipoprotein receptors induced by evolocumab treatment, these two factors may underlie the massive reduction of VLDL particles. 
+On the other hand, apo(a) can be recycled to the extracellular space and bind to apoB-containing particles again, resulting in moderate reduction of Lp(a) [@Sharma2017]. 
+We find that our hypothesis has connections to other studies. 
+First, in the ANITSCHKOW trail, evolocumab had no impact on arterial wall inflammation in patients with persistent Lp(a) elevation [@Stiekema2018]. 
+This could be caused by the fact that inhibition of PCSK9 may increase other receptors such as VLDL receptors and CD36. 
+Both the VLDL receptor and CD36 are present in macrophages and endothelial cells. 
+The accelerated uptake of a Lp(a)-TRL complex via these receptors would not attenuate foam cell formation. 
+Second, when Lp(a) is directly targeted via antisense oligonucleotide treatment, reduction of Lp(a) production does decrease cellular inflammatory responses [@Viney2016]. 
+In this case the reduced Lp(a) production may decrease the abundance of Lp(a)-TRL complex and uptake of TRL by macrophages, thereby reducing foam cell formation. 
+Future studies are required in order to test our hypothesis.
 
@@
 
-Strengths and limitations of our study warrant consideration. 
 Our study provided the first NMR metabolomics data for a PCSK9 inhibition trial. 
-The detailed metabolomic profiling revealed metabolic effects of the PCSK9 inhibitor, and improved our understanding of this lipid-lowering therapy. 
-Moreover, in this study we developed a statistical method to separate Lp(a) content from the LDL measurements. 
-However, due to the heterogeneity of Lp(a) and limited number of participants, our estimation of the proportions of LDL with apo(a) had very large uncertainty. 
-Applying our statistical model with a larger number of participants as well as validation in the laboratory are warranted in the future studies.
+The detailed metabolomic profiling not only allowed us to reveal systemic effects of PCSK9 inhibition, but also to assess the relationship between Lp(a) lowering and 14 lipoprotein subclasses. 
+We did not detect statistically significant correlations between Lp(a) lowering and reduction in lipoprotein subclasses probably due to the small number of participants and considerable missing data in some lipoprotein subclasses.
+We feel it is worth to apply the multilevel multivariate model developed in this study in a larger clinical trial to improve our understanding of interrelationship between Lp(a) lowering and metabolism of other lipoprotein subclasses.
 
-In conclusion, our NMR metabolomic profiling revealed that Lp(a) reduction by PCSK9 inhibition corresponded to substantial reduction in VLDL, IDL and LDL particles as well as their lipid contents. 
-Our analysis suggested that alternative pathways beyond LDL receptor mediated clearance may be involved in Lp(a) reduction by PCSK9 inhibition.
+In conclusion, our NMR metabolomic profiling revealed that evolocumab treatment in patients with high Lp(a) massively decreases VLDLs in addition to the well known effect on LDLs. 
+The substantial reduction in VLDLs after evolocumab treatment suggested enhanced clearance of VLDL particles in patients with high Lp(a).
 

### 2020-01-16 manuscript.Rmd
introduction with comments from Erik Stroes

@@
 Background: 
-Epidemiological studies show that subjects with elevated lipoprotein(a) [Lp(a)] have a higher risk for developing cardiovascular diseases. 
-Inhibition of proprotein convertase subtilisin/kexin type 9 (PCSK9) is one of the limited options to lower Lp(a). 
-However, effects of PCSK9 inhibition with evolocumab on circulating metabolites such as lipoprotein subclasses, amino acids and fatty acids remain to be characterized.
+Epidemiological studies substantiated that subjects with elevated lipoprotein(a) [Lp(a)] have a markedly increased cardiovascular risk. 
+Inhibition of proprotein convertase subtilisin/kexin type 9 (PCSK9) lowers both LDL cholesterol (LDL-C) as well as Lp(a), albeit modestly. 
+Effects of PCSK9 inhibition on circulating metabolites such as lipoprotein subclasses, amino acids and fatty acids remain to be characterized.
 
 Methods:
-In this study, we performed nuclear magnetic resonance (NMR) metabolomics on plasma samples derived from 30 individuals (placebo N = 14; evolocumab N = 16). 
+We performed nuclear magnetic resonance (NMR) metabolomics on plasma samples derived from 30 individuals (placebo N = 14; evolocumab N = 16). 
 We assessed the effects of evolocumab treatment on circulating metabolites by running lognormal regression analyses. 
@@
 On average, evolocumab treatment for 16 weeks resulted in a 17% (95% credible interval: 8% to 26%) reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
-Although, we did not observe correlations between Lp(a) lowering and reduction in other 14 lipoprotein subclasses, we identified that increasing concentrations of baseline Lp(a) were associated with larger reduction in triglyceride-rich VLDL particles after evolocumab treatment.
+Although we did not observe correlations between Lp(a) lowering and reduction in other 14 lipoprotein subclasses, we identified that increasing concentrations of baseline Lp(a) were associated with larger reduction in triglyceride-rich VLDL particles after evolocumab treatment.
 
 Conclusions:
-Inhibition of PCSK9 with evolocumab massively decreased VLDL in addition to the well known effect on LDL. 
+Inhibition of PCSK9 with evolocumab markedly reduced VLDL particle concentrations in addition to lowering LDL-C. 
 The substantial reduction in VLDLs after evolocumab treatment indicates enhanced clearance of VLDL particles in patients with high Lp(a).
@@
 
-lipoprotein(a) [Lp(a)], synthesized in the liver, is a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail. 
-Similar to other lipoproteins, Lp(a) is cleared from the circulation primarily by the liver [@Cain2005]. 
-Epidemiological studies showed that subjects with elevated Lp(a) had increasing risk of developing cardiovascular disease [@ODonoghue2019]. 
-The biological mechanism underlying the cardiovascular risk of Lp(a) remains to be elucidated. 
-One mechanism is that Lp(a) may induce macrophage apoptosis and plaque necrosis through activation of a pathway involving CD36 (Cluster of differentiation 36) and TLR2 (Toll-like receptor 2) [@Seimon2010].
-Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce Lp(a) in plasma [@VanCapelleveen2016]. 
-One of the agents that are capable to lower circulating Lp(a) levels is proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody, which prevents extracellular PCSK9 from binding to the LDL receptor, and results in supraphysiological abundance of LDL receptors [@Raal2014; @Raal2016].
-The excess number of LDL receptors on the surface of hepatocytes were hypothesized to be responsible for the enhanced Lp(a) catabolism [@Romagnuolo2015; @Raal2016]. 
-However, the above hypothesis does not concur with the effect of statin therapy, which significantly increases hepatic LDL receptor abundances but also increases plasma Lp(a) levels [@Tsimikas2019]. 
-The same hypothesis was also challenged by the observation that LDL receptor knock out mice had similar Lp(a) catabolism compared to wild type mice [@Cain2005]. 
+Lipoprotein(a) [Lp(a)], synthesized in the liver, is a low-density lipoprotein (LDL)-like particle covalently bound to the apolipoprotein(a) [apo(a)]. 
+Lp(a) is cleared from the circulation primarily by the liver [@Cain2005], and the exact mechanism remains to be elucidated. 
+Epidemiological studies substantiated a markedly increased cardiovascular risk in subjects with elevated Lp(a), involving both pro-inflammatory as well as pro-coagulant effects [@ODonoghue2019]. 
+To date, limited therapeutic options are available to reduce Lp(a) in plasma [@VanCapelleveen2016]. 
+One agent that lowers Lp(a) levels is proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody, resulting in increased abundance of LDL receptors [@Raal2014; @Raal2016].
+The excess number of LDL receptors on the surface of hepatocytes may contribute to enhanced Lp(a) catabolism [@Romagnuolo2015; @Raal2016]. 
+However, this concept is not supported the effect of statin therapy, which also increases hepatic LDL receptor abundances with a concomitant increase of plasma Lp(a) [@Tsimikas2019]. 
 Hence, the exact role of the LDL-receptor in Lp(a) catabolism has not been elucidated and remains a matter of debate [@Reyes-Soffer2017; @Watts2018; @Stoekenbroek2018].
-The LDL receptor is not the only target of PCSK9. 
-PCSK9 also induces the degradation of the VLDL (very low density lipoprotein) receptor, ApoER2 (apolipoprotein E receptor 2), and CD36 [@Poirier2008; @Demers2015]. 
-Interestingly, the VLDL receptor and CD36 are receptors for triglyceride-rich lipoproteins (TRLs) as well as Lp(a), indicating a possible interrelationship between Lp(a) and TRL metabolism [@Argraves1997; @Seimon2010; @Dijk2018].
-The metabolic link between Lp(a) and TRL metabolism was confirmed by a recent NMR (nuclear magnetic resonance) metabolomics study, in which a Lp(a)-raising genotype was identified to associate with decreasing concentrations of triglyceride-rich VLDL particles [@Kettunen2016].
-However, the mechanism underlying this inverse relationship is unknown. 
-As a cost-effective approach, NMR metabolomics simultaneously quantifies hundreds of circulating biomarkers and provides insights on multiple metabolic pathways [@Soininen2009].
-A NMR metabolomics study showed that genetic inhibition of PCSK9 and statin therapy had similar effects on circulating metabolites [@Sliz2018]. 
-However, genetic inhibition of PCSK9 is different from inhibition with monoclonal antibody since genetic inhibition includes intracellular functions of PCSK9. 
-Furthermore, although PCSK9 monoclonal antibody is known for lowering LDL cholesterol and Lp(a), the relationship between Lp(a) lowering and metabolism of other lipoproteins has not been studied.  
-In the current study, we analyzed NMR metabolomics of plasma derived from 30 subjects with elevated Lp(a) (placebo 14; evolocumab 16). 
-We first ran lognormal regression analyses to identify metabolic changes corresponding to evolocumab treatment.  
-Next, we developed a multilevel multivariate model to assess the relationship between Lp(a) lowering and metabolism of 14 lipoprotein subclasses.
+
+Besides the LDL receptor, PCSK9 also induces the degradation of the VLDL (very low density lipoprotein) receptor, ApoER2 (apolipoprotein E receptor 2), and CD36 (cluster of differentiation 36) [@Poirier2008; @Demers2015].
+Interestingly, the VLDL receptor and CD36 are receptors for triglyceride-rich lipoproteins (TRLs) as well as Lp(a), indicating a potential interaction between Lp(a) and TRL metabolism [@Argraves1997; @Seimon2010; @Dijk2018].
+A link between Lp(a) and TRL metabolism is supported by a recent NMR (nuclear magnetic resonance) metabolomics study, in which a Lp(a)-raising genotype was found to associate with decreasing concentrations of triglyceride-rich VLDL particles [@Kettunen2016].
+The mechanism underlying this inverse relationship is unknown. 
+Another NMR metabolomics study showed that genetic inhibition of PCSK9 and statin therapy had similar effects on circulating metabolites [@Sliz2018]. 
+However, genetic inhibition of PCSK9 differs from inhibition with monoclonal antibody, since genetic inhibition includes intracellular functions of PCSK9. 
+Moreover, data on the relationship between Lp(a) lowering and metabolism of other lipoproteins are scarce.
+
+In the present study, we analyzed NMR metabolomics of plasma derived from 30 subjects with elevated Lp(a), using either placebo (n = 14) or PCSK9 inhibitor (evolocumab, n = 16).
 
@@
 
-This was a site-specific sub study of the ANITSCHKOW trial (NCT02729025) conducted at the Amsterdam UMC in The Netherlands between April 2016 and July 2017. 
+This was a substudy of the ANITSCHKOW trial (NCT02729025) conducted at the Amsterdam UMC in The Netherlands between April 2016 and July 2017. 
 The ANITSCHKOW trial was a phase 3b, multicenter, randomized, double-blind, placebo-controlled trial using subcutaneous injections of evolocumab 420mg Q4W for 16 weeks as investigational product [@Stiekema2018].

### 2020-01-16 manuscript.Rmd
Results with comments of Erik Stroes

@@
 
-In the present study, we analyzed NMR metabolomics of plasma derived from 30 subjects with elevated Lp(a), using either placebo (n = 14) or PCSK9 inhibitor (evolocumab, n = 16).
+In the present study, we studied 30 subjects with elevated Lp(a), using either placebo (n = 14) or PCSK9 inhibitor (evolocumab, n = 16). 
+In particular, we analyzed NMR metabolomics of plasma derived from these participants at baseline and 16 weeks after treatment of evolocumab.
 
@@
 
-The current study included 30 patients (evolocumab, n = 16; placebo, n = 14). 
-Baseline characteristics of the two groups were comparable and are provided in Table 1.
+Baseline characteristics of the evolocumab and the placebo group were comparable (Table 1).
 
@@
 
-However, we identified that the reduction in medium-sized VLDL particles was associated with increasing baseline Lp(a) concentrations (Pearson correlation coefficient -0.5 [-0.8, -0.06]). 
+Interestingly, we observed that the reduction in medium-sized VLDL particles was associated with increasing baseline Lp(a) concentrations (Pearson correlation coefficient -0.5 [-0.8, -0.06]). 
 The correlations between baseline Lp(a) concentrations and reduction in other VLDL particles, including extremely large, very large, large, small and very small VLDLs, were -0.1 [-0.6, 0.3], -0.03 [-0.4, 0.4], -0.3 [-0.7, 0.2], -0.1 [-0.7, 0.6], and 0.2 [-0.3, 0.7], respectively.

### 2020-01-17 manuscript.Rmd
Discussion after comments from Erik Stroes

@@
 Inhibition of PCSK9 with evolocumab markedly reduced VLDL particle concentrations in addition to lowering LDL-C. 
-The substantial reduction in VLDLs after evolocumab treatment indicates enhanced clearance of VLDL particles in patients with high Lp(a).
+The extent of reduction in VLDL particles depended on the baseline level of Lp(a). 
+Our findings suggest a marked effect of evolocumab on VLDL metabolism in subjects with elevated Lp(a).
 
@@
 
-In the present study, we studied 30 subjects with elevated Lp(a), using either placebo (n = 14) or PCSK9 inhibitor (evolocumab, n = 16). 
-In particular, we analyzed NMR metabolomics of plasma derived from these participants at baseline and 16 weeks after treatment of evolocumab.
+In the present study, we aim to study metabolic effects of PCSK9 inhibition with evolocumab in subjects with elevated Lp(a). 
+To this end, we studied 30 subjects with elevated Lp(a), using either placebo (n = 14) or PCSK9 inhibitor (evolocumab, n = 16). 
+We analyzed NMR metabolomics of plasma derived from these participants at baseline and 16 weeks after treatment of evolocumab. 
+We found that evolocumab had a marked effect on VLDL metabolism in subjects with elevated Lp(a).
 
@@
 
+## Metabolic effects of PCSK9 inhibition with evolocumab
+
 Baseline characteristics of the evolocumab and the placebo group were comparable (Table 1).
-
-## Metabolic effects of PCSK9 inhibition with evolocumab
 
@@
 
-The major finding of this work is that evolocumab treatment in patients with high Lp(a) massively decreases VLDLs in addition to the well known effect on LDLs. 
-In the current work, evolocumab treatment resulted in on average 17% reduction in plasma Lp(a) concentration, together with 67% and 21% reduction in LDL cholesterol and triglyceride. 
-These treatment effects were similar to the outcomes reported in the past, indicating that the subjects involved in the current study were representative for the entire ANITSCHKOW trail [@Stiekema2018].
+Here, we report that evolocumab markedly decreases VLDL particle concentrations in addition to lowering LDL-C in patients with elevated Lp(a). 
+In particular, we identified that the extent of VLDL reduction was dependent on the baseline Lp(a) level, suggesting a marked effect of evolocumab on VLDL metabolism in subjects with elevated Lp(a).
 
-Our detailed metabolomic profiling revealed that PCSK9 inhibition with evolocumab resulted in a substantial reduction in concentrations of VLDL, IDL and LDL particles as well as their lipid and fatty acid contents.
-We did not observe changes in other circulating metabolites such as amino acids and ketone bodies. 
-In general, PCSK9 inhibition with evolocumab resulted in similar metabolic changes as genetic inhibition of *PCSK9* and statin therapy [@Wurtz2016; @Sliz2018]. 
-However, in contrast to genetic inhibition of *PCSK9* and statin therapy that had much larger effects on LDL particles than VLDL particles, we observed that evolocumab treatment induced a more substantial reduction in VLDL particles compared to LDL particles.
-Our multivariate analysis showed that Lp(a) lowering did not correlate with reduction in any of the 14 lipoprotein subclasses, suggesting alternative pathways in Lp(a) catabolism, such as scavenger receptor class B type I (SR-BI) [@Yang2013] and plasminogen receptor [@Sharma2017].
+Overall evolocumab resulted in a 17% reduction of Lp(a), a 67% reduction of LDL-C and a 21% reduction of triglycerides. 
+Metabolomic consequences of PCSK9 inhibition with evolocumab were similar to those induced by genetic inhibition of *PCSK9* and statin therapy [@Wurtz2016; @Sliz2018]. 
+The major metabolic effects of PCSK9 inhibition were on lipoproteins and their lipid contents.  
+There is little effects on other circulating metabolites such as amino acids and ketone bodies. 
+In contrast to genetic inhibition of *PCSK9* and statin therapy that had much larger effects on LDL particles than VLDL particles, we observed that evolocumab treatment induced a more substantial reduction in VLDL particles compared to LDL particles.
+Although we observed substantial reduction in extremely and very large VLDL particle concentrations, the reduction of triglycerides was modest. 
+This is because the particle concentrations of these two VLDL particles were very low, and the VLDLs of medium and small size were the main carriers of triglycerides (Figure S1). 
+With the help of detailed NMR metabolomics, we were able to visualize the changes in VLDL particles even though the overall change of triglycerides was modest. 
+However, since this study focused on the subjects with elevated Lp(a), future studies are required to elucidate if evolocumab has similar effect on VLDL particles in subjects without elevated Lp(a).
+
+![Figure S1. Triglycerides in lipoprotein subclasses](./Triglycerides_in_lipoproteins.png)
+
+Our multivariate analysis showed that Lp(a) lowering did not correlate with particle concentration reduction in any of the 14 lipoprotein subclasses, suggesting alternative pathways for Lp(a) clearance, such as scavenger receptor class B type I (SR-BI) [@Yang2013] or the plasminogen receptor [@Sharma2017].
 
@@
 We observed that subjects with higher baseline Lp(a) also showed tendency to have a larger reduction in extremely large and large VLDL particle concentrations after treatment with evolocumab. 
-However, those tendencies had large uncertainty due to considerable number of missing observations. 
+However, those tendencies had large uncertainty due to considerable number of missing observations caused by very low concentrations that were below the detection limit (Figure S2).
+
+![Figure S2 Particle concentrations of VLDLs.](./VLDL_particle_concentrations.png)
+
 The percentage of missing observations in extremely large and large were 69% and 47%, respectively. 
@@
 In contrast, the percentage of missing observations in medium VLDL was 9%, allowing us to assess the relationship between its reduction and baseline Lp(a) levels. 
-On the other hand, we did not observe similar relationship between baseline Lp(a) and reduction in small and very small VLDL particles, although the percentages of missing observations were small for these two lipoprotein subclasses (6% and 3%), suggesting that Lp(a) may not influence catabolism of these two triglyceride-poor VLDL particles.
+On the other hand, we did not observe similar relationship between baseline Lp(a) and reduction in small and very small VLDL particles, although the percentages of missing observations were small for these two lipoprotein subclasses (6% and 3%), suggesting that Lp(a) may not influence catabolism of these two VLDL particles.
 Our observation is in line with the association between the Lp(a)-raising genotype and decreasing concentrations of extremely large, very large, large and medium VLDL particles, but not small and very small VLDL particles [@Kettunen2016].
@@
 On the other hand, apo(a) can be recycled to the extracellular space and bind to apoB-containing particles again, resulting in moderate reduction of Lp(a) [@Sharma2017].
-We find that our hypothesis has connections to other studies. 
-First, in the ANITSCHKOW trail, evolocumab had no impact on arterial wall inflammation in patients with persistent Lp(a) elevation [@Stiekema2018]. 
-This could be caused by the fact that inhibition of PCSK9 may increase other receptors such as VLDL receptors and CD36. 
-Both the VLDL receptor and CD36 are present in macrophages and endothelial cells. 
-The accelerated uptake of a Lp(a)-TRL complex via these receptors would not attenuate foam cell formation. 
-Second, when Lp(a) is directly targeted via antisense oligonucleotide treatment, reduction of Lp(a) production does decrease cellular inflammatory responses [@Viney2016]. 
-In this case the reduced Lp(a) production may decrease the abundance of Lp(a)-TRL complex and uptake of TRL by macrophages, thereby reducing foam cell formation. 
+Based on the above, we hypothesize that the large effect of evolocumab on VLDL particles in subjects with elevated Lp(a) is due to abundant Lp(a)-TRL complex in these patients. 
 Future studies are required in order to test our hypothesis.
@@
 
-In conclusion, our NMR metabolomic profiling revealed that evolocumab treatment in patients with high Lp(a) massively decreases VLDLs in addition to the well known effect on LDLs. 
-The substantial reduction in VLDLs after evolocumab treatment suggested enhanced clearance of VLDL particles in patients with high Lp(a).
+In conclusion, our NMR metabolomic profiling revealed that evolocumab treatment in patients with high Lp(a) markedly decreases VLDLs in addition to the well known effect on LDLs. 
+Moreover, we found that the extent of VLDL reduction was dependent on the baseline Lp(a) level. 
+Our findings suggest a marked effect of evolocumab on VLDL metabolism in subjects with elevated Lp(a).
 

### 2020-05-11 univariate_lognormal_regression_analysis.R
Univariate lognormal regression analyses for assessing evolocumab effects on lipoprotein subclasses and Lp(a)

@@ -1,0 +1,348 @@
+#
+# This script dscribed the univariate lognormal regression analyses
+# for lipoportein subclasses and LP(a), comparing evolocumab and 
+# placebo group
+#
+library(readr)
+library(dplyr)
+library(tidyr)
+library(tibble)
+library(rstan)
+library(ggplot2)
+library(multidplyr)
+library(purrr)
+library(broom)
+library(tidybayes)
+library(stringr)
+library(GGally)
+library(forcats)
+
+############################################
+# Import data
+############################################
+
+# load metabolomics data and clinical data
+Nightingale <- read_rds("Anitschkow_metabolomics.rds")
+Anitschkow <- read_rds("Anischkow_meta.rds")
+
+
+Anitschkow <- Anitschkow %>% 
+  rename(Treatment = `Randomized treatment Group`)
+
+sample_visit <- Anitschkow %>% 
+  select(sampleid:Treatment, Statine_use, Ezetimibe) %>% 
+  distinct()
+
+############################################
+# 1. Analysis-ready data preparation
+############################################
+
+# reshpae clinical data (LP[a], LDLc and TG) for
+# lognormal regression model for assessing the effects
+# of evolocumab
+anitschkow_biochemical_dat <- function(df, which_biochemical_compound) {
+  df %>% 
+    filter(what == which_biochemical_compound) %>% 
+    select(patient, visit, Treatment, Statine_use, Ezetimibe, what, y) %>% 
+    spread(visit, y) %>% 
+    mutate(v1_mis = ifelse(is.na(V1), 1L, 0L), 
+           v1_cen = ifelse(V1 == 0, 1L, 0L), 
+           v2_mis = ifelse(is.na(V2), 1L, 0L), 
+           v2_cen = ifelse(V2 == 0, 1L, 0L))
+}
+
+d.lpa <- anitschkow_biochemical_dat(Anitschkow, "Lipoprotein(a) nmol/l")
+d.ldlc <- anitschkow_biochemical_dat(Anitschkow, "LDL cholesterol mmol/l")
+d.tg <- anitschkow_biochemical_dat(Anitschkow, "Triglyceride mmol/l")
+
[…]

### 2020-05-11 multivariate_multi_level_regression_analysis.R
multivariate multi level model for assessing relationship between Lp(a) reduction and lipoprotein subclasses changes

@@ -1,0 +1,165 @@
+#
+# This script descirbed the multivariate multi level model
+# which is used to evaluate relationship between Lp(a)
+# reduction and other lipoprotein subclasses changes
+#
+library(readr)
+library(dplyr)
+library(tidyr)
+library(tibble)
+library(rstan)
+library(ggplot2)
+library(multidplyr)
+library(purrr)
+library(broom)
+library(tidybayes)
+library(stringr)
+library(forcats)
+
+# load the multivariate multi level model
+AK3 <- stan_model("multivariate_multi_level_regression_model.stan")
+
+#######################################################
+# 1. Import data
+#######################################################
+
+# load metabolomics data and clinical data
+Nightingale <- read_rds("Anitschkow_metabolomics.rds")
+Anitschkow <- read_rds("Anischkow_meta.rds")
+
+Anitschkow <- Anitschkow %>% 
+  rename(Treatment = `Randomized treatment Group`)
+
+########################################################
+# 2. Data preparation
+########################################################
+
+dat.lpa <- Anitschkow %>% 
+  filter(what == "Lipoprotein(a) nmol/l", Treatment == "AMG145") %>% 
+  select(-what, -unit) %>% 
+  rename(LPa = y)
+
+Nightingale.nested <- dat.lpa %>% 
+  left_join(Nightingale, by = "sampleid") %>% 
+  select(patient, visit, Treatment, LPa, what, y, miss, cens) %>% 
+  mutate(y = y * 10^9) %>% 
+  group_by(what) %>% 
+  nest() %>% 
+  filter(what %in% c("XXL-VLDL-P", "XL-VLDL-P", 'L-VLDL-P', "M-VLDL-P", "S-VLDL-P", "XS-VLDL-P", 
+                     "IDL-P", "L-LDL-P", 'M-LDL-P', "S-LDL-P", 
+                     "XL-HDL-P", "L-HDL-P", "M-HDL-P", "S-HDL-P"))
+
+# convert the nested lipoprotein data into stan friendly data
+anitschkow_dat <- function(my_dat) {
+  subj_dat <- my_dat %>% 
+    select(patient, Treatment) %>% 
+    distinct()
+  
+  N <- nrow(my_dat)
+  L <- my_dat$LPa
+  N_mis <- sum(my_dat$miss)
[…]

### 2020-05-11 multivariate_multi_level_regression_analysis.R
small adjust

@@ -16,9 +16,6 @@ library(tidybayes)
 library(stringr)
 library(forcats)
 
-# load the multivariate multi level model
-AK3 <- stan_model("multivariate_multi_level_regression_model.stan")
-
 #######################################################
 # 1. Import data
 #######################################################
@@ -91,6 +88,9 @@ lipoproteins <- Nightingale.nested %>%
 ##########################################################
 # 3. Run multivariate multi level regression analyses
 ##########################################################
+
+# load the multivariate multi level model
+AK3 <- stan_model("multivariate_multi_level_regression_model.stan")
 
 xiang_cluster <- create_cluster(cores = 10) %>% 
   cluster_library("purrr") %>% 

### 2020-05-11 univariate_lognormal_regression_model_adj_lipid_lowering.R
univariate lognormal regression model adjusting for lipid lowering drug usage

@@ -1,0 +1,183 @@
+#
+# This script describes univariate lognormal regression analyses for 
+# Lp(a) and lipoprotein subclasses particle concentrations, adjusting
+# usage of lipid lowering drugs
+#
+library(readr)
+library(dplyr)
+library(tidyr)
+library(tibble)
+library(rstan)
+library(ggplot2)
+library(multidplyr)
+library(purrr)
+library(broom)
+library(tidybayes)
+library(stringr)
+
+##########################################################
+# 1. Load data
+##########################################################
+
+Nightingale <- read_rds("Anitschkow_metabolomics.rds")
+Anitschkow <- read_rds("Anischkow_meta.rds")
+
+Nightingale1 <- Nightingale %>% 
+  select(sampleid:y)
+
+Anitschkow1 <- Anitschkow %>% 
+  rename(Treatment = `Randomized treatment Group`) %>% 
+  select(sampleid:Ezetimibe) %>% 
+  distinct()
+
+############################################################
+# 2. Data preparation
+############################################################
+
+Nightingale.nested <- Anitschkow1 %>% 
+  left_join(Nightingale1, by = "sampleid") %>% 
+  select(-sampleid) %>% 
+  spread(visit, y) %>% 
+  group_by(what) %>% 
+  nest()
+
+anitschkow_dat <- function(my_dat) {
+  
+  # convert from mol/L to nmol/L (lipoprotein particle concentration)
+  y1 <- my_dat$V1 * 10^9
+  y2 <- my_dat$V2 * 10^9
+  
+  # missing observations
+  which_y1_mis <- which(is.na(y1))
+  N_y1_mis <- length(which_y1_mis)
+  
+  which_y2_mis <- which(is.na(y2))
+  N_y2_mis <- length(which_y2_mis)
+  
+  # observations equal to 0
+  which_y1_cen <- which(y1 == 0)
+  N_y1_cen <- length(which_y1_cen)
+  y1[which_y1_cen] <- NA
+  
+  which_y2_cen <- which(y2 == 0)
+  N_y2_cen <- length(which_y2_cen)
+  y2[which_y2_cen] <- NA
+  
+  # observed
+  which_y1_obs <- which(!is.na(y1))
+  N_y1_obs <- length(which_y1_obs)
+  
+  which_y2_obs <- which(!is.na(y2))
[…]

### 2020-05-11 manuscript.Rmd
Adjustments according to reviewers

@@
 ---
-title: "Metabolic Effects of PCSK9 Inhibition With Evolocumab in Subjects with Elevated Lp(a)"
 output: word_document
 bibliography: /home/xiang/Documents/Bibliography/lpa.bib
-csl: journal-of-lipid-research.csl
+csl: lipids-in-health-and-disease.csl
 ---
 
-Xiang Zhang, PhD$^1,^2$,
-Lotte C.A. Stiekema, MD$^2$,
-Erik S.G. Stroes, PhD$^2$,
-Albert K. Groen, PhD$^1$
+Metabolic Effects of PCSK9 Inhibition With Evolocumab in Subjects with Elevated Lp(a)
 
-1. Department of Experimental Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
-2. Human and Animal Physiology, Wageningen University, De Elst 1, 6708 WD Wageningen, The Netherlands
-3. Department of Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
+Xiang Zhang, xiang.zhang@wur.nl Department of Experimental Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands; Human and Animal Physiology, Wageningen University, De Elst 1, 6708 WD Wageningen, The Netherlands
+
+Lotte C.A. Stiekema, l.c.stiekema@amsterdamumc.nl Department of Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
+
+Erik S.G. Stroes, e.s.stroes@amsterdamumc.nl Department of Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
+
+Albert K.
+Groen, a.k.groen@amsterdamumc.nl Department of Experimental Vascular Medicine, Amsterdam University Medical Center, University of Amsterdam, Meibergdreef 9, 1105 AZ Amsterdam, The Netherlands
 
@@
 Methods:
-We performed nuclear magnetic resonance (NMR) metabolomics on plasma samples derived from 30 individuals (placebo N = 14; evolocumab N = 16). 
-We assessed the effects of evolocumab treatment on circulating metabolites by running lognormal regression analyses. 
-Subsequently, we assessed the interrelationship between Lp(a) and 14 lipoprotein subclasses in response to treatment of evolocumab, by running multilevel multivariate regression analyses.
+We performed nuclear magnetic resonance (NMR) metabolomics on plasma samples derived from 30 individuals with elevated Lp(a) (> 150 mg/dL). 
+The 30 participants were randomly assigned into two groups, placebo (N = 14) and evolocumab (N = 16). 
+We assessed the effect of 16 weeks of evolocumab 420mg Q4W treatment on circulating metabolites by running lognormal regression analyses, and compared this to placebo. 
+Subsequently, we assessed the interrelationship between Lp(a) and 14 lipoprotein subclasses in response to treatment with evolocumab, by running multilevel multivariate regression analyses.
 
 Results: 
-On average, evolocumab treatment for 16 weeks resulted in a 17% (95% credible interval: 8% to 26%) reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
-Although we did not observe correlations between Lp(a) lowering and reduction in other 14 lipoprotein subclasses, we identified that increasing concentrations of baseline Lp(a) were associated with larger reduction in triglyceride-rich VLDL particles after evolocumab treatment.
+On average, evolocumab treatment for 16 weeks resulted in a 17% (95% credible interval: 8% to 26%, P < 0.001) reduction of circulating Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles as well as their lipid contents.
+Interestingly, increasing concentrations of baseline Lp(a) were associated with larger reduction in triglyceride-rich VLDL particles after evolocumab treatment.
 
@@
 
-# Introduction
+Trial registration:
+Clinical trial registration information is registered at ClinicalTrials.gov on April 14, 2016 with the registration number NCT02729025.
+
+Keywords: PCSK9 antibodies; Evolocumab; Lipoprotein(a); Metabolomics; VLDL
+
+# Background
 
 Lipoprotein(a) [Lp(a)], synthesized in the liver, is a low-density lipoprotein (LDL)-like particle covalently bound to the apolipoprotein(a) [apo(a)]. 
-Lp(a) is cleared from the circulation primarily by the liver [@Cain2005], and the exact mechanism remains to be elucidated. 
+Lp(a) is cleared from the circulation primarily by the liver [@Cain2005], but the exact mechanism remains to be elucidated. 
 Epidemiological studies substantiated a markedly increased cardiovascular risk in subjects with elevated Lp(a), involving both pro-inflammatory as well as pro-coagulant effects [@ODonoghue2019]. 
@@
 The excess number of LDL receptors on the surface of hepatocytes may contribute to enhanced Lp(a) catabolism [@Romagnuolo2015; @Raal2016]. 
-However, this concept is not supported the effect of statin therapy, which also increases hepatic LDL receptor abundances with a concomitant increase of plasma Lp(a) [@Tsimikas2019]. 
-Hence, the exact role of the LDL-receptor in Lp(a) catabolism has not been elucidated and remains a matter of debate [@Reyes-Soffer2017; @Watts2018; @Stoekenbroek2018].
+However, this concept is not supported by the effect of statin therapy, which also increases abundance of the hepatic LDL receptor with a concomitant increase of plasma Lp(a) [@Tsimikas2019]. 
+Hence, the exact role of the LDL-receptor in Lp(a) catabolism remains a matter of debate [@Reyes-Soffer2017; @Watts2018; @Stoekenbroek2018].
 
@@
 Another NMR metabolomics study showed that genetic inhibition of PCSK9 and statin therapy had similar effects on circulating metabolites [@Sliz2018]. 
-However, genetic inhibition of PCSK9 differs from inhibition with monoclonal antibody, since genetic inhibition includes intracellular functions of PCSK9. 
+However, genetic inhibition of PCSK9 differs from inhibition with a monoclonal antibody, since genetic inhibition includes intracellular functions of PCSK9. 
 Moreover, data on the relationship between Lp(a) lowering and metabolism of other lipoproteins are scarce.
 
-In the present study, we aim to study metabolic effects of PCSK9 inhibition with evolocumab in subjects with elevated Lp(a). 
-To this end, we studied 30 subjects with elevated Lp(a), using either placebo (n = 14) or PCSK9 inhibitor (evolocumab, n = 16). 
-We analyzed NMR metabolomics of plasma derived from these participants at baseline and 16 weeks after treatment of evolocumab. 
+The aim of the current study was to investigate metabolic effects of PCSK9 inhibition (evolocumab) in patients with elevated Lp(a). 
+To this end, we studied 30 subjects with elevated Lp(a), using either placebo (n = 14) or the PCSK9 inhibitor (evolocumab, n = 16). 
+We analyzed NMR metabolomics of plasma derived from these participants at baseline and 16 weeks after treatment. 
 We found that evolocumab had a marked effect on VLDL metabolism in subjects with elevated Lp(a).
@@
 The ANITSCHKOW trial was a phase 3b, multicenter, randomized, double-blind, placebo-controlled trial using subcutaneous injections of evolocumab 420mg Q4W for 16 weeks as investigational product [@Stiekema2018].
-The study was conducted in accordance to the principles of the Declaration of Helsinki and the study protocol was approved by the ethic committee of the Amsterdam UMC.
+The current study was based on 30 patients (placebo N = 14, evolocumab N = 16). 
+Written informed consent was obtained from each patient included in the study. 
+The study protocol conforms to the ethical guidelines of the Declaration of Helsinki. 
+The study protocol was approved by the ethic committee of the Amsterdam UMC.
 
-Patients were found to be eligible for enrollment in the ANITSCHKOW trial when they were $\geq$ 50 years of age, had a fasting LDL-C of $\geq$ 2.6 mmol/L (100mg/dL), and an Lp(a) of $\geq$ 125nmol/L (50mg/dL).
-At the time of randomization, study subjects who were on lipid lowering treatment had to be on a stable dosing regimen for at least 8 weeks. 
-Exclusion criteria were among others, diabetes mellitus and a cardiovascular event within three months prior to randomization. 
-The complete list of eligibility criteria was reported previously [@Stiekema2018].
+The complete list of eligibility criteria for the ANITSCHKOW trial were:
+
+* Inclusion criteria
+
+  + Patient has provided written informed consent prior to initiation of any study specific activities/procedures
+
+  + Male or female, $\geq$ 50 years of age at the time of informed consent
+
+  + Fasting Lp(a) $\geq$ 125 nmol/L (50 mg/dL)
+
+  + Fasting LDL-C $\geq$ 2.6 mmol/L (100 mg/dL)
+
+  + For patients receiving lipid-lowering therapy (not required to participate in this study), lipid-lowering therapy, including statin dose, must be unchanged for $\geq$ 8 weeks prior to screening TBRmax above 1.6 (either right carotid, left carotid or thoracic aorta) on FDG-PET/CT
+
+* Exclusion criteria
+
+  + Currently receiving, or < 4 weeks since receiving, treatment in another investigational device or drug study(ies), or participating in other investigational procedures
+
+  + Known diagnosis of diabetes mellitus or screening fasting serum glucose $\geq$ 7 mmol/L or glycated haemoglobin (HbA1c) $\geq$ 6.5%
+
+  + History of homozygous familial hypercholesterolemia
+
+  + Recent cardiovascular event (myocardial infarction, unstable angina, percutaneous coronary intervention [PCI], coronary artery bypass graft, or stroke) within 3 months prior to randomization, or planned cardiac surgery, PCI or carotid stenting, or planned major non-cardiac surgery during the course of the study period
+
+  + Currently undergoing lipid apheresis
+
+  + Known contraindications or limitations to FDG-PET/CT (eg, scanner weight limit, devices that can cause image artifacts, or carotid/aortic stents/grafts)
+
+  + Autoimmune disease/vasculitis, active inflammatory diseases, proven or suspected bacterial infections
+
+  + Recent (< 1 month prior to screening) or ongoing serious infection requiring intravenous antibiotic therapy
+
+  + Recent (< 6 weeks prior to screening) or current treatment with medications that may have a significant effect on plaque inflammation as measured by plaque TBR, including: oral, rectal, or injectable corticosteroids or immunosuppressive medications (eg, cyclosporine, methotrexate, tacrolimus, azathioprine, anti-thymocyte globulin, sirolimus, anti-tumour necrosis factor agents such as infliximab, anti-interleukin [IL] 6 therapy such as tocilizumab, or anti-IL1 therapy)
+
+  + Recent (< 6 weeks prior to screening) or current treatment with aspirin (> 325 mg/day) or nonsteroidal anti-inflammatory drugs (NSAIDs) (>1000 mg/day)
+
+  + Known sensitivity to any of the active substances or excipients (eg, carboxymethylcellulose) to be administered during dosing
+
+  + Treatment with a cholesterol ester transfer protein inhibitor (eg, anacetrapib, dalcetrapib, evacetrapib) or mipomersen or lomitapide in the last 12 months prior to screening
+
+  + Known systemic disorders such as hepatic, renal, hematologic, and malignant diseases or any clinically significant medical condition that could interfere with the conduct of the study
+
+  + History of malignancy (except non-melanoma skin cancers, cervical in-situ carcinoma, breast ductal carcinoma in situ, or stage 1 prostate carcinoma) within the last 5 years
+
+  + Patients likely to not be available to complete all protocol-required study visits or procedures, or unreliability as a study participant (eg, alcohol or other drug abuse in the past year or psychosis), to the best of the patient’s and investigator’s knowledge
+
+  + History or evidence of any other clinically significant disorder, condition or disease that, in the opinion of the investigator or sponsor physician, if consulted, would pose a risk to subject safety or interfere with the study evaluation, procedures, or completion
+
+  + Prior treatment with evolocumab or any other therapy to inhibit PCSK9
+
+  + Pregnant or breastfeeding or planning to become pregnant or breastfeed during treatment with study drug and for an additional 15 weeks after the last dose of study drug
 
@@
 The following 14 lipoprotein subclasses and their lipid compositions were quantified: 
-extremely large (average particle diameter >75 nm), 
+extremely large (average particle diameter > 75 nm), 
 very large (average particle diameter 64.0 nm), large (53.6 nm), medium (44.5 nm), small (36.8 nm) and very small VLDL (31.3 nm); 
@@
 2) metabolite concentration at time of randomization ($x$). 
+To investigate potential combined effects of evolocumab and other lipid lowering drugs on lipoproteine subclasses, we added two indicator variables $S$ and $E$.
+If a patient was treated with statin, then $S = 1$, otherwise $S = 0$.
+Similarly, if a patient was also treated with ezetimibe, then $E = 1$, otherwise $E = 0$.
 The baseline concentrations were centered and scaled so that the mean is 0 and standard deviation is 1.
@@
 
-On average, evolocumab treatment for 16 weeks resulted in a 17% (95% credible interval: 8% to 26%) reduction in Lp(a), together with a concomitant 67% (57%, 76%) and 21% (6% 35%) reduction in LDL cholesterol and triglyceride, respectively.
+On average, evolocumab treatment for 16 weeks resulted in a 17% (95% credible interval: [8%, 26%]) reduction in Lp(a), together with a concomitant 67% [57%, 76%] and 21% [6%, 35%] reduction in LDL cholesterol and triglyceride, respectively.
 
 To identify the metabolic effects corresponding to PCSK9 inhibition with evolocumab, we performed NMR metabolomics covering metabolic pathways such as lipoprotein subclasses, fatty acids, amino acids and glycolysis.
-We observed that evolocumab treatment resulted in substantial reduction in particle concentration of extremely large (80% [48% 100%]), very large (90% [70% 100%]), large (60% [34%, 83%]), medium (50% [36%, 63%]), small (39% [32% 46%]) and very small VLDL (47% [40% 53%]).
-We also observed that evolocumab treatment resulted in particle concentration reduction in IDL (53% [45%, 60%]), large (56% [48% 65%]), medium (59% [50%, 67%]) and small LDL (55% [47%, 64%]). 
-In addition, we observed that evolocumab treatment resulted in decreased concentrations of very large HDL particles (24% [3% 45%]), and increased concentrations of medium HDL particles (13% [4% 23%]) (Figure 1).
+We observed that evolocumab treatment resulted in substantial reduction in particle concentration of extremely large (80% [48%, 100%]), very large (90% [70%, 100%]), large (60% [34%, 83%]), medium (50% [36%, 63%]), small (39% [32%, 46%]) and very small VLDL (47% [40%, 53%]).
+We also observed that evolocumab treatment resulted in particle concentration reduction in IDL (53% [45%, 60%]), large (56% [48%, 65%]), medium (59% [50%, 67%]) and small LDL (55% [47%, 64%]). 
+In addition, we observed that evolocumab treatment resulted in decreased concentrations of very large HDL particles (24% [3%, 45%]), and increased concentrations of medium HDL particles (13% [4%, 23%]) (Figure 1).
 
-![Figure 1 Mean difference between evolocumab and palacebo group, adjusting for pre-treatment lipoprotein particle concentrations.
-Circles represent the posterior means.
-Lines refer to the 95% credible intervals](./Treatment_effect_lipoproteins.png)
+Interestingly, we observed that evolocumab treatment had similar effects on lipoprotein subclasses in subjects with or without usage of lipid lowering medication.
+Compared to patients treated by statin, we observed that evolocumab treatment resulted in similar but larger effects on lipoproteins in subjects with treatment of both statin and ezetimibe (Figure S1).
 
-Similar to the lipoprotein particle concentration profiles, we observed that evolocumab treatment resulted in substantial reduction of esterified cholesterol (CE), triglyceride (TG), free cholesterol (FC) and phospholipid (PL) in VLDL, IDL and LDL as well as very large HDL (Figure 2).
-
-![Figure 2 Mean difference in lipoprotein lipid compositions between evolocumab and placebo group, adjusting for pre-treatment lipoprotein lipid concentrations.
-Circles represent the posterior means.
-Lines refer to the 95% credible intervals.
-CE: esterified cholesterol; FC: free cholesterol; PL: phospholipids; TG: triglyceride.](./Treatment_effect_lipoprotein_lipid.png)
-
+Similar to the lipoprotein particle concentration profiles, we observed that evolocumab treatment resulted in substantial reduction of esterified cholesterol (CE), triglyceride (TG), free cholesterol (FC) and phospholipid (PL) in VLDL, IDL and LDL as well as very large HDL (Figure S2).
 The NMR metabolomics also quantified the fatty acid content in the lipoproteins. 
-We observed that treatment of evolocumab resulted in 30% [24% 36%] reduction in total fatty acids, with the largest effect on docosahexaenoic acid (DHA 40% [24% 55%]) (Figure 3).
-
-![Figure 3 Mean difference in fatty acid concentrations between evolocumab and placebo group, adjusting for pre-treatment fatty acid concentrations.
-Circles represent the posterior means.
-Lines refer to the 95% credible intervals.](./Treatment_effect_FFAs.png)
-
+We observed that treatment of evolocumab resulted in 30% [24%, 36%] reduction in total fatty acids, with the largest effect on docosahexaenoic acid (DHA 40% [24%, 55%]) (Figure S3).
 We observed no difference in concentrations of other metabolites including amino acids, fluid balance, glycolysis and ketone bodies.
@@
 To identify the relationship between Lp(a) lowering and reduction in lipoprotein subclasses, we developed a multilevel multivariate model. 
-We observed that Lp(a) lowering was not associated with reduction of the 14 lipoprotein subclasses (Figure 4).
+We observed that Lp(a) lowering was not associated with reduction of the 14 lipoprotein subclasses (Figure S4). 
+Interestingly, we observed that the reduction in medium-sized VLDL particles was associated with increasing baseline Lp(a) concentrations (Pearson correlation coefficient -0.5 [-0.8, -0.06]) (Figure 2).
 
-![Figure 4 Relationship between Lp(a) lowering and reduction in 14 lipoprotein subclasses.
-Circles represent the posterior mean Pearson correlation coefficients.
-Lines refer to the 95% credible intervals.](./corr_lipoproteins_lpa.png)
-
-Interestingly, we observed that the reduction in medium-sized VLDL particles was associated with increasing baseline Lp(a) concentrations (Pearson correlation coefficient -0.5 [-0.8, -0.06]). 
 The correlations between baseline Lp(a) concentrations and reduction in other VLDL particles, including extremely large, very large, large, small and very small VLDLs, were -0.1 [-0.6, 0.3], -0.03 [-0.4, 0.4], -0.3 [-0.7, 0.2], -0.1 [-0.7, 0.6], and 0.2 [-0.3, 0.7], respectively.
-
-![Figure 5 Reduction of medium VLDL particles correlated with baseline lipoprotein(a) concentrations.
-Every circle represents the posterior mean reduction of medium VLDL particle concentration and the posterior mean of baseline lipoprotein(a) in a patient treated with evolocumab.
-The verticle and horizontal bar represents the 95% credible interval.
-The blue dashed line represented the average percentage (50%) change in medium VLDL particle after evolocumab treatment.](./Baseline_lpa_mVLDL.png)
 
@@
 Metabolomic consequences of PCSK9 inhibition with evolocumab were similar to those induced by genetic inhibition of *PCSK9* and statin therapy [@Wurtz2016; @Sliz2018]. 
-The major metabolic effects of PCSK9 inhibition were on lipoproteins and their lipid contents.  
-There is little effects on other circulating metabolites such as amino acids and ketone bodies. 
+The major metabolic effects of PCSK9 inhibition were on apoB containing lipoproteins and their lipid contents. 
+We observed that evolocumab treatment resulted in decreasing concentration of very large HDL particles and increasing concentrations of medium sized HDL particles. 
+Although our observation is consistent with a recent study of evolocumab on lipoprotein particles, there is no evidence suggesting that PCSK9 modulates HDL particle production or clearance [@Toth2018].
+Future studies are needed to investigate the role of PCSK9 in HDL metabolism. 
+There was little effect on other circulating metabolites such as amino acids and ketone bodies. 
 In contrast to genetic inhibition of *PCSK9* and statin therapy that had much larger effects on LDL particles than VLDL particles, we observed that evolocumab treatment induced a more substantial reduction in VLDL particles compared to LDL particles.
 Although we observed substantial reduction in extremely and very large VLDL particle concentrations, the reduction of triglycerides was modest. 
-This is because the particle concentrations of these two VLDL particles were very low, and the VLDLs of medium and small size were the main carriers of triglycerides (Figure S1). 
+This is because the absolute concentrations of triglycerides in these two VLDL particles were very low, and the VLDLs of medium and small size were the main carriers of triglycerides (Figure S5). 
 With the help of detailed NMR metabolomics, we were able to visualize the changes in VLDL particles even though the overall change of triglycerides was modest. 
 However, since this study focused on the subjects with elevated Lp(a), future studies are required to elucidate if evolocumab has similar effect on VLDL particles in subjects without elevated Lp(a). 
-
-![Figure S1. Triglycerides in lipoprotein subclasses](./Triglycerides_in_lipoproteins.png)
-
 Our multivariate analysis showed that Lp(a) lowering did not correlate with particle concentration reduction in any of the 14 lipoprotein subclasses, suggesting alternative pathways for Lp(a) clearance, such as scavenger receptor class B type I (SR-BI) [@Yang2013] or the plasminogen receptor [@Sharma2017].
-
 Interestingly, we identified that subjects with higher baseline Lp(a) showed a larger reduction in medium VLDL particle concentrations after treatment with evolocumab. 
 We observed that subjects with higher baseline Lp(a) also showed tendency to have a larger reduction in extremely large and large VLDL particle concentrations after treatment with evolocumab. 
-However, those tendencies had large uncertainty due to considerable number of missing observations caused by very low concentrations that were below the detection limit (Figure S2).
-
-![Figure S2 Particle concentrations of VLDLs.](./VLDL_particle_concentrations.png)
-
-The percentage of missing observations in extremely large and large were 69% and 47%, respectively. 
-Similarly, since the very large VLDL particle had the largest percentage of missing data (72%) among all the 14 lipoprotein subclasses, we simply had not enough data to detect the relationship between its reduction and baseline Lp(a).
-In contrast, the percentage of missing observations in medium VLDL was 9%, allowing us to assess the relationship between its reduction and baseline Lp(a) levels. 
-On the other hand, we did not observe similar relationship between baseline Lp(a) and reduction in small and very small VLDL particles, although the percentages of missing observations were small for these two lipoprotein subclasses (6% and 3%), suggesting that Lp(a) may not influence catabolism of these two VLDL particles.
+However, those tendencies had large uncertainty due to the fact that the number of these VLDL particles present after treatment was below the detection limit, precluding assessment of correlations (Figure S6).
+The number of medium VLDL particles was sufficiently high to accurately assess the relationship between its reduction and baseline Lp(a) levels. 
+Surprisingly, we did not observe a similar relationship between baseline Lp(a) and reduction in small and very small VLDL particles, suggesting that Lp(a) may not influence catabolism of these two VLDL particles.
 Our observation is in line with the association between the Lp(a)-raising genotype and decreasing concentrations of extremely large, very large, large and medium VLDL particles, but not small and very small VLDL particles [@Kettunen2016].
@@
 We feel it is worth to apply the multilevel multivariate model developed in this study in a larger clinical trial to improve our understanding of interrelationship between Lp(a) lowering and metabolism of other lipoprotein subclasses.
+Another limitation of this study is that we exclusively focused on the patients with elevated Lp(a) ($\geq$ 125nmol/L), in a future study it will be interesting to see if evolocumab treatment will result in similar effects on triglyceride-rich VLDL particles in patients with less elevated Lp(a) (for example $\geq$ 30 mg/dL).
+
+# Conclusions
 
@@
 
-# Authors
+# List of abbreviations
 
-# Acknowledgements
+Lp(a): lipoprotein(a)
+PCSK9: proprotein convertase subtilisin/kexin type 9
+LDL-C: LDL cholesterol
+NMR: nuclear magnetic resonance
+VVLVLDL: extremely large VLDL
+VLVLDL: very large VLDL
+LVLDL: large VLDL 
+MVLDL: medium VLDL
+SVLDL: small VLDL 
+VSVLDL: very small VLDL
+IDL: intermediate density lipoprotein
+LLDL: large LDL
+MLDL: medium LDL
+SLDL: small LDL 
+VLHDL: very large HDL
+LHDL: large HDL 
+MHDL medium HDL
+SHDL: small HDL
 
-# Sources of Funding
+# Declarations
 
-# Disclosures
+## Ethics approval and consent to participate
 
-# Affiliations
+Written informed consent was obtained from each patient included in the study. 
+The study protocol conforms to the ethical guidelines of the Declaration of Helsinki. 
+The study protocol was approved by the ethic committee of the Amsterdam UMC.
+
+## Consent for publication
+
+Not applicable'
+
+## Availability of data and materials
+
+The datasets used and/or analysed during the current study are available from the corresponding author on reasonable request.
+
+## Competing interests
+
+E.S.G.S. reports that his institution has received lecturing fees and advisory board fees from Amgen Inc., Regeneron, Sanofi, Akcea, Novartis and Athera. 
+All other authors declared no conflict of interest.
+
+## Funding
+
+This ANITSCHKOW trial was funded by Amgen Inc.
+
+## Authors' contributions
+
+XZ performed the analysis and composed the manuscript draft. 
+L.C.A.S. prepared the data.
+E.S.G.S. supervised the project and edited the manuscript. 
+A.K.G. supervised the project and edited the manuscript.
+
+## Acknowledgements
+
+Not applicable
 
